Missense mutations in CRX homeodomain cause dominant retinopathies through two distinct mechanisms

Version 1

Abstract

Summary

Homeodomain transcription factors (HD TFs) are instrumental to vertebrate development. Mutations in HD TFs have been linked to human diseases, but their pathogenic mechanisms remain elusive. Here we use Cone-Rod Homeobox (CRX) as a model to decipher the disease-causing mechanisms of two HD mutations, p.E80A and p.K88N, that produce severe dominant retinopathies. Through integrated analysis of molecular and functional evidence in vitro and in knock-in mouse models, we uncover two novel gain-of-function mechanisms: p.E80A increases CRX-mediated transactivation of canonical CRX target genes in developing photoreceptors; p.K88N alters CRX DNA-binding specificity resulting in binding at ectopic sites and severe perturbation of CRX target gene expression. Both mechanisms produce novel retinal morphological defects and hinder photoreceptor maturation distinct from loss-of-function models. This study reveals the distinct roles of E80 and K88 residues in CRX HD regulatory functions and emphasizes the importance of transcriptional precision in normal development.

Introduction

Homeodomain transcription factors (HD TFs) play a fundamental role in vertebrate development. Members of the HD TF family are characterized by the presence of a highly conserved 60 amino acid helix-turn-helix DNA-binding domain known as the homeodomain (HD). The HD is one of the most studied eukaryotic DNA-binding motifs since its discovery in Drosophila homeotic transformations1,2. Hundreds of HD TFs have subsequently been documented in regulating gene expression programs important for body plan specification, pattern formation and cell fate determination1,2. Mutations in HD TFs have been linked to many human diseases, including neuropsychiatric and neurodegenerative conditions3,4. Although significant progress has been made in understanding HD-DNA interactions, uncovering the pathogenetic mechanisms of disease-causing missense mutations in HD have proven challenging.

The retina has long been used as a model system to study the role of HD TFs in normal central nervous system (CNS) development and in neurological diseases5. During retinogenesis, HD TFs play essential roles in the patterning of neuroepithelium, specification of retinal progenitors and differentiation of all retinal cell classes that derive from a common progenitor6. Importantly, many HD TFs are shared between the brain and the retina during development and mutations in these TFs can lead to disease manifestation in both tissues7–12. The accessibility and wealth of available molecular tools makes the retina a valuable tool to decipher the pathogenic mechanisms of HD TF mutations associated with neurological diseases.

Here, we study CRX, a HD TF essential for photoreceptor cells in the retina, as a model to understand how single amino acid substitutions in the HD impact TF functions and cause blinding diseases. Photoreceptors are the most numerous neurons in the retina and are specialized to sense light and initiate vision through a process called phototransduction. Animal studies have demonstrated that Crx is first expressed in post-mitotic photoreceptor precursors and maintained throughout life13,14. Loss of CRX results in impaired photoreceptor gene expression, failure of maturation and rapid degeneration of immature, non-functional photoreceptors15. Protein-coding sequence variants in human CRX have been associated with inherited retinal diseases (IRDs) that affect photoreceptors: Leber congenital amaurosis (LCA), cone-rod dystrophy (CoRD), and retinitis pigmentosa (RP) (OMIM:602225). However, the recessive phenotype observed in Crx knockout mouse models fails to recapitulate many dominant human CRX mutations that arise de novo15.

CRX contains two functional domains – the N-terminal HD and C-terminal activation domain (AD) (Figure 1A), both are required for proper activation of target genes and maintenance of normal Crx mRNA transcript abundance16,17. To understand how CRX HD mutations cause diseases, we have previously reported a mutation knock-in mouse model carrying a hypomorphic mutation p.R90W (R90W) in CRX HD18,19. We found that R90W mutation produces a recessive loss-of-function phenotype very similar to that of Crx knockout mice.

Disease associated missense mutations altered CRX HD DNA binding specificity.

(A) Diagram of CRX functional domains: Homeodomain (HD) for DNA-binding and Activation Domain (AD) for target gene transactivation. The three missense mutations in this study are located at the C-terminus of CRX HD and associated with different retinal diseases in human. Number in the parenthesis denotes the CRX HD position of the corresponding mutated residue. (B) Alignments of HD recognition helix sequences for the indicated HD proteins for which HD missense mutations have been associated with inherited diseases. Accession numbers can be found in Supplementary Table S1. Missense variants in this study (highlighted) are located at highly conserved residues across species and different HD TFs. (C) Spec-seq experimental workflow (STAR Methods). (D) Spec-seq library design of monomeric HD binding sites. (E) EMSA gel images of Spec-seq experiments with different CRX HD species. Bx: Bound. B-: Unbound. (F) Relative binding energy comparison from two different experiments with WT HD. (G) Binding energy model for WT CRX HD. (H-J) Relative binding energy comparison between WT HD and R90W HD (H), E80A HD (I), or K88N HD (J). Consensus sequence is defined to have relative binding energy of 0kT (TAATCC for WT, R90W and E80A, TAATTA for K88N). The diagonal line of identity is represented in grey dash. The orange dashed line shows the best linear fit to the data. (K-M) Binding energy models for R90W HD (K), E80A HD (L), and K88N HD (M). Only sequence variants within two mismatches to the corresponding consensus sequences were used to generate binding models. Negative binding energy is plotted such that bases above the x-axis are preferred bases and bases below the x-axis are unfavorable bases. Constant bases (TAA) carried no information are drawn at arbitrary height in grey.

Intriguingly, several missense mutations within the same HD recognition helix as R90W, including p.E80A (E80A) and p.K88N (K88N), are linked to severe dominant IRDs20–22 (Figures 1A and 1B). Here, we utilized a multi-omics approach to investigate the functional consequences of the E80A and K88N mutations on CRX regulatory activities and photoreceptor development (Figure S1). Comparison of the in vitro HD-DNA binding models of CRX and disease variants generated by Spec-seq revealed unique specificity changes of each mutant protein. Introduction of each mutation into the endogenous Crx locus generated knock-in mouse models CrxE80A and CrxK88N that reproduced dCoRD- and dLCA-like phenotypes. ChIP-seq analysis of CRX binding in vivo revealed mutation-specific changes in CRX targetome, consistent with mutation-specific DNA binding changes in vitro. Retinal RNA-seq analysis uncovered two distinct mechanisms by which the two HD missense mutations contribute to altered gene expression programs during photoreceptor differentiation and maturation. Our results highlight the importance of residues E80 and K88 in CRX-mediated transcriptional regulation during photoreceptor development and the diverse mechanisms by which HD missense mutations can affect TF functions and lead to severe dominant neurological diseases.

Results

K88N but not E80A mutation alters CRX HD DNA-binding specificity in vitro

CRX belongs to the paired-like HD TF family that recognize a 6-bp DNA motif in a stereotypic way23–29. Extensive studies of the HD have revealed a canonical HD-DNA recognition model where recognition of the 3’ region (bases 4-6) of the HD DNA binding site is mediated by specificity determinants within the conserved HD recognition helix, corresponding to CRX residues 80-9623–26,29,30 (Figure 1B). In particular, HD residue 50, equivalent to CRX K88 residue (Figure 1A), is the specificity determinant in paired-like HD TF-DNA interactions24,25. Since the three disease-associated HD missense mutations, E80A, K88N and R90W, are located within the CRX HD recognition helix, we wondered if these mutations change CRX HD DNA-binding specificity.

We adapted a high-throughput in vitro assay, Spec-seq, that determines protein-DNA-binding specificity by sequencing31–33. Spec-seq was developed based on the traditional electrophoretic mobility shift assay (EMSA) to measure protein-DNA interactions. Spec-seq allows us to measure the relative binding affinities (i.e., specificity) for a library of HD binding motifs in parallel and generate quantitative binding models for different CRX HDs (Figures 1C). Based on the HD-DNA interaction model, we designed and tested a Spec-seq library containing all possible monomeric HD motifs (TAANNN) (Figure 1D).

We first obtained the wild-type (WT) CRX HD DNA binding model with Spec-seq using bacterially-expressed and affinity-purified HD peptides (Figures 1E-1G, STAR Methods). Relative binding energies of CRX WT HD from two experiments showed strong correlation (r: 0.984) and noise level (0.114 kT) within the expected range in typical Spec-seq data (Figure 1F). Binding energy model of WT HD was then generated by applying multiple linear regressions on the relative binding energies of all sequences within two base-pair mismatches to the WT CRX consensus (TAATCC)13 (Figure 1G, STAR Methods). A clear preference for CC bases at the 3’ end of the motif is consistent with known CRX binding preference in vitro and in vivo34,35.

We next sought to understand how disease mutations affect CRX DNA-binding specificity. We purified all mutant HD peptides following the same protocol as WT HD peptides and verified their DNA binding (Figures S2A-S2D)17. Comparison of the relative binding energies between each pair of mutant and WT HD revealed distinct effects (Figures 1H-1J). By definition, the consensus DNA binding motif of a testing peptide has a relative binding energy of 0kT. The relative binding energy difference between nucleotide variants and the consensus motif correlates with the DNA binding specificity of the testing peptide. We found that R90W HD and E80A HD both prefer the same consensus motif as WT (Figures 1K and 1L). R90W HD bound with slightly higher specificity than WT, as demonstrated by most data points falling above the diagonal line (Figures 1H and 1K), suggesting that R90W HD is more sensitive than WT to binding sequence variations. In contrast, when comparing E80A HD with WT HD, many data points fell below the diagonal line and the relative binding energies regressed towards 0 on the E80A axis (Figures 1I and 1L). This suggests that E80A HD bound with lower specificity than WT HD and thus was more tolerant to base variations in the HD DNA motif. Different from R90W and E80A, K88N mutation dramatically altered CRX HD DNA-binding specificity (r: 0.160) (Figures 1J and 1M). The K88N preferred binding sequence (TAAT/ATT/A) is referred to as N88 HD motif hereafter. K88N HD also had the largest degree of discrimination from its preferred to the weakest binding motif, suggesting that it is most sensitive to variants in the HD DNA motif. As a control, we tested a second library with the TAANNN sites on the reverse strand and obtained similar results (Figures S2E-S2L). Together, these results indicate that while E80A mutation does not affect CRX HD DNA binding specificity, the K88N mutation dramatically alters the specificity in vitro.

E80A protein binds to WT sites while K88N occupies novel genomic regions with N88 HD motifs in vivo

Next, we asked if changes in DNA-binding specificity affected mutant CRX chromatin binding in developing photoreceptors. We first created two human mutation knock-in mouse models, CrxE80A and CrxK88N, each carrying a single base substitution at the endogenous Crx locus (Figures S3A and S3B, STAR Methods). We confirmed that Crx mRNA was expressed at comparable levels in WT and mutant retinas (Figure S3C), and the full-length CRX proteins were readily detectable in the nuclear extracts from all samples (Figure S3D). We then obtained genome-wide binding profiles for each CRX variant by chromatin immunoprecipitation followed by sequencing (ChIP-seq) on mouse retinas at P14, a time when all retinal cell types are born, photoreceptor specification is completed in WT animals, and prior to any observed cell death in other CRX mutants previously characterized18,36. To focus on changes specific to each mutant CRX protein, only homozygous animals were used for ChIP-seq profiling.

Unsupervised clustering of all CRX binding sites revealed two major clusters (Figure 2A, STAR Methods). Cluster 1 consisted of canonical WT CRX binding sites that are also occupied by CRX E80A protein (Figures 2A and 2B). Similar to WT CRX, CRX E80A binding in vivo was mostly enriched in intronic, followed by intergenic and TSS regions (Figure 3D). In contrast, CRX R90W, that also showed similar consensus preference to WT in vitro, failed to produce significant DNA binding in vivo (Figures 2A-2C, STAR Methods). This suggests that the retinopathy phenotype of CrxR90W/W is likely due to loss of binding at canonical WT CRX binding sites. Intriguingly, while CRX K88N showed loss of binding at canonical CRX binding sites, it gained a small set of binding sites (Cluster 2, Figures 2A-2C). De novo motif searching with DREME37 under CRX peaks in each genotype revealed enrichment of monomeric HD motifs (Figure 2E) consistent with those found in Spec-seq (Figures 1G and 1K-1M), highlighted by a change in enriched HD motif from WT CRX HD type to N88 HD type in the CrxK88N/N retinas. Consistency with in vitro binding models suggests that in vivo changes in CRX chromatin binding were at least in part driven by the intrinsic changes in HD-DNA binding specificity by each individual mutation.

CRX E80A binds to WT sites while CRX K88N occupies novel genomic regions enriched for N88 HD motif in vivo.

(A) Enrichment heat map depicting CRX ChIP-seq normalized reads centered at all possible CRX peaks ± 2kb, sorted by binding intensity in WT samples. Clusters were defined by hierarchical clustering of CRX binding intensity matrix from all genotypes (STAR Methods). (B-C) Genome browser representations of ChIP-seq normalized reads for different CRX species in P14 WT and mutant mouse retinas at Rho and Atf2. (D) Enrichment heatmap showing fraction of CRX ChIP-seq peaks fall in different genomic environments. (E) Logo representations of de novo found short HD motifs under CRX ChIP-seq peaks in WT and mutant mouse retinas with DREME E-value on the right.

CRX-dependent activated genes affected in opposite directions in developing CrxE80A and CrxK88N mutant retinas

(A) Heat map showing sample-wise Pearson correlations of the expression of all CRX-dependent activated genes between P10 WT and HD mutant mouse retinas in this study (rows) with post-natal WT retinas from age P3 to P21 (columns, data from GSE87064). (B) Heat map showing the expression changes of DEGs in CRX-dependent activated gene set in HD mutant mouse retinas at P10. (C-D) Heat maps showing expression changes of selected photoreceptor genes from Group 1 and Group 2. Color scale identical to (B).

E80A and K88N mutations affected the expression of CRX-dependent activated genes in opposite directions in a critical time window of photoreceptor differentiation

To understand how different CRX mutations affected gene expression at canonical and de novo binding sites and how these changes impair photoreceptor differentiation, we turned to bulk RNA-seq analysis from the developing retinas at P10. At P10, photoreceptors have started to differentiate, and the expression of many photoreceptor genes undergo exponential increase38,39. To focus on the most relevant expression changes, we first defined a set of genes that most likely depend on CRX activity nearby for expression (Figure S4A, STAR Methods). Briefly, we associated each CRX ChIP-seq peak to the nearest gene, filtered only genes with nearby CRX ChIP-seq peaks, and further narrowed the list of genes to those significantly down-regulated in the loss of function mutant CrxR90W/W. Gene ontology (GO) analysis confirmed that this putative CRX-dependent gene set is associated with biological processes related to photoreceptor development and functions (Figure S4B). This set of putative CRX-dependent genes also showed developmental dependent gain in expression, consistent with CRX’s primary function as a transcriptional activator (Figure S4C). As a control, CRX-independent genes were constitutively expressed and largely involved in general cellular processes (Figures S4D-S4F). Therefore, the CRX-dependent gene set comprises genes important for photoreceptor differentiation and functional maturation and are dependent on CRX for activation. We denote these genes as “CRX-dependent activated genes” (CRX-DAGs).

Next, we sought to understand how each mutation affected photoreceptor differentiation. One way of measuring the progression of photoreceptor differentiation is to determine the similarity in CRX-DAG expression in a given sample with that of known developmental ages in WT control animals. We thus performed sample-wise correlation of CRX-DAG expression obtained in our P10 samples with a previously published RNA-seq dataset of normal mouse retinal development (Figure 3A)38. As expected, our P10 WT sample showed strong correlations with all developmental ages in the published WT control dataset. A stronger correlation with early ages (P3, P7, P10) and a weaker correlation with later ages (P14, P21) is also an indication of ongoing photoreceptor differentiation at P10. Unlike the WT sample, CrxE80A/+ and CrxE80A/A samples both showed a stronger correlation with later developmental ages (P14, P21) but a weaker correlation with earlier postnatal ages (P3, P7). Since the CRX-DAGs are normally developmentally upregulated, this shift toward correlation with later developmental ages suggested that these genes were prematurely upregulated in the P10 CrxE80A mutant retinas. In contrast, CrxK88N/+ and CrxK88N/N samples both showed a weaker correlation with all developmental ages when compared with WT samples in our dataset. This suggests that early photoreceptor differentiation was compromised in both CrxK88N mutants, consistent with their association with early-onset LCA22. Importantly, CrxR90W/W, also associated with LCA-like phenotype18,21, displayed strong correlation with earlier ages (P3, P7) similar to WT, but only showed moderate correlation with later ages (P14, P21). This suggests loss of CRX function at canonical binding sites does not affect the initiation of photoreceptor differentiation, but WT CRX activity at these sites is required to sustain differentiation. Since CRX-DAG expression was more severely affected in CrxK88N mutants than in CrxR90W/W, the photoreceptor differentiation deficits seen in the CrxK88N mutants cannot be explained solely by the loss of regulatory activity at canonical CRX binding sites. Overall, our sample-wise correlation analysis with normal retinal development dataset suggests that E80A and K88N mutations affected the expression CRX-dependent activated genes in opposite directions, implicating novel and distinct pathogenic mechanisms from the loss-of-function R90W mutation.

CrxE80A retinas show up-regulation of rod genes but down-regulation of cone genes, underlying CoRD-like phenotype

Upon closer examination, we noted that not all CRX-DAGs were up-regulated in CrxE80A mutants (Figure 3B). Hierarchical clustering of all CRX-DAGs using expression changes revealed two major groups (STAR Methods). In aggregate, when compared to WT, Group 1 genes were up-regulated in CrxE80A mutants while Group 2 genes were down-regulated. We noted genes indicative of the two photoreceptor subtypes, rods and cones, could partially define the two groups (Figures 3C and 3D). For example, Esrrb40 and Nrl41 in Group 1 are important regulators of rod differentiation. Other genes in Group 1 are components of the phototransduction cascade in rods, including Rcvrn42, Rho43, Gnat144,45, Pde6g46, Abca447,48, Gnb149, Rdh1250, Cngb151, Rp152,53. Mis-regulations of many of these genes have been associated with diseases that affect rod development, function, and long-term survival. The increased activation of these genes likely underlies the stronger correlation with later developmental ages in CrxE80A+/ and CrxE80A/A retinas (Figure 3A). In contrast, Group 2 genes, many down-regulated in CrxE80A mutants, were implicated in cone development and functions. For example, Gnat254,55, Pde6c56, and Pde6h57,58 all act in the cone phototransduction cascade. Mis-regulation of these genes has also been implicated in different retinal dystrophies that primarily affect cone photoreceptors. Comparison of ChIP-seq signal revealed that peaks associated with Group 2 genes showed lower occupancy compared to Group 1 genes in the CrxE80A/A retinas (Mann-Whitney U test p-value: 9.51e-07) while no difference was observed in WT retinas (Mann-Whitney U test p-value: 0.541), suggesting loss of CRX activity likely underlies the down-regulation of Group 2 genes in the CrxE80A mutants (Figure S5A). Collectively, the selective down-regulation of cone genes in Group 2 may explain the CoRD-like phenotype in adult CrxE80A mutant mice described later.

Additionally, we noticed that a subset of genes not affected in CrxR90W/W (CRX-independent genes) were also down-regulated in CrxE80A mutants (Figures S5B-S5D). Among these genes were transcription regulators important for early photoreceptor development, such as Ascl159, Rax60, Sall361, and Pias362. The down-regulation of these factors coincided with the up-regulation of mature rod genes in P10 CrxE80A retinas, suggesting that the E80A mutation might hamper the proper timing of photoreceptor differentiation.

CrxK88N retinas display greater reduction of rod and cone genes than the loss-of-function mutants

CrxK88N retinas had the most severe gene expression changes among all mutants with down-regulation of both Group 1 (rod) and Group 2 (cone) genes (Figures 3B-3D). The heterozygous CrxK88N/+ retina displayed a similar degree of expression reduction as homozygous CrxR90W/W, consistent with its association with dLCA22. Given the normal phenotype of heterozygous loss-of-function mutants -Crx+/- and CrxR90W/+18, these results suggested that mutant CRX K88N not only failed to activate WT target genes, but also functionally antagonized WT CRX regulatory activity in differentiating photoreceptors. This antagonism might be associated with ectopic CRX K88N activity when bound to regulatory regions with N88 HD DNA motifs (Figures 2A and 2E). In the absence of WT CRX, CrxK88N/N retina displayed a more severe expression reduction of CRX-DAG than CrxR90W/W, raising the possibility that CRX K88N also antagonized the activity of other transcriptional regulators important for photoreceptor differentiation. Supporting this possibility, a set of CRX-independent genes were also mis-regulated in both heterozygous and homozygous CrxK88N mutants, we noted that a number of these down-regulated genes are also involved in photoreceptor functional development (Figures S5B, S5E, and S5F). Overall, CRX K88N is associated with greater gene expression changes than other CRX HD mutants, which is likely attributed to ectopic regulatory activity.

Both E80A and K88N mutants show compromised rod/cone terminal differentiation in young adults

Since Crx+/- and CrxR90W/+ mutant mouse models showed a late-time recovery in photoreceptor gene expression and function19, we sought to determine the degree of photoreceptor differentiation in CrxE80A and CrxK88N mutants at P21. At this age, the normal retina has largely completed terminal differentiation with photoreceptor gene expression reaching a plateau63. However, when P21 CRX HD mutant retinas were examined for the expression of genes under the GO term detection of light stimulus (GO:0009583), which comprises genes in both rod and cone phototransduction cascades, many genes failed to reach WT levels, despite variable degrees of impact across different HD mutants (Figure 4A). CrxE80A mutants, in contrast to the increased rod gene expression at P10, displayed a deficiency in both cone and rod phototransduction genes at P21 (Figure 4B, Supplementary Table 6). This suggests that premature gene mutant CRX E80A transcriptional activity fails to sustain photoreceptor terminal differentiation and ultimately results in non-functional and severely affected photoreceptors. In comparison, CrxK88N mutants showed severely reduced expression of rod/cone phototransduction genes at both P10 and P21 (Figures 3B-3D and 4B).

Photoreceptor genes important for phototransduction are down-regulated in all HD mutants

(A) Box plot showing that genes in the detection of light stimulus GO term were down-regulated and affected to various degrees in different adult (P21) HD mutant mouse retinas. (B) Heat map showing that expression of both cone and rod phototransduction genes were down-regulated in adult (P21) HD mutant mouse retinas. Annotation of rod and cone enrichment of each gene is in Supplementary Table 6. See Supplementary Figure S6 for the developmental expression dynamics of these genes.

To assess retinal morphology and photoreceptor subtype-specific gene expression at the cellular level, we performed immunohistochemistry analysis on P21 retinal sections. In WT animals, a hallmark of photoreceptor maturation is the outgrowth of photoreceptor outer segments (OS) filled with proteins necessary for the phototransduction. We thus performed hematoxylin and eosin (H&E) staining on P21 sagittal retinal sections to visualize changes in retinal layer organization, focusing on photoreceptor layers – outer nuclear layer (ONL) and OS. Compared to the well-organized ONL in WT retinas (Figure 5A), all mutants showed variable degrees of ONL disorganization, forming waves, whorls, and rosettes (Figures 5B-5E). The ONL disorganization was more severe in homozygotes than in heterozygotes for both mutations and CrxK88N mutants were more severely affected than CrxE80A mutants. Photoreceptor OS layer was formed in the CrxE80A/+ retinas, but absent in CrxE80A/A, CrxK88N/+, and CrxK88N/N mutant retinas. Inner retinal layers, including inner plexiform layer (IPL) and ganglion cell layer (GCL) were not as severely affected as the outer retinal layers, supporting a model that the mutant morphological abnormalities largely originated from the diseased photoreceptors. These morphological abnormalities were distinct from the degenerative phenotypes of other Crx mutant models reported previously18,64,65.

Only CrxE80A/+ retinas maintain photoreceptor OS and residual rod ERG response

(A-E) Hematoxylin-eosin (H&E) staining of P21 retina sections show that photoreceptor OS layer is absent in all mutant retinas except CrxE80A/+. OS: outer segment; ONL: outer nuclear layer; INL: inner nuclear layer; IPL: inner plexiform layer; GCL: ganglion cell layer. Scale bar, 100μm. (F-J) Rhodopsin (RHO, red) immunostaining is present in CrxE80A/+, CrxE80A/A, and CrxK88N/+ retinas and absent in CrxK88N/N retina. Cone arrestin (mCAR, green) immunostaining is absent in all mutant retinas. Nuclei were visualized by DAPI staining (Blue). Scale bar, 100μm. (K-M) The electroretinogram responses (ERG) recorded from 1-month mice. Error bars represent the standard error of the mean (SEM, n ≥ 4). p-value: Two-way ANOVA and Tukey’s multiple comparisons. ****: p ≤ 0.0001. ns: >0.05.

Next, we sought to determine the expression of the rod-specific visual pigment rhodopsin (RHO) and cone arrestin (mCAR) in the P21 mouse retinas. In WT retinas, RHO is trafficked to the rod OS while mCAR is present in the cone OS and IS (inner segment), cell body, and synaptic terminals (Figure 5F). Unlike WT retina, all mutants lacked mCAR immunoreactivity (Figures 5G-5J), consistent with the loss of cone gene expression shown by RNAseq (Figure 4B). In CrxE80A/+ retinas, RHO staining was localized to the OS layer; in CrxE80A/A and CrxK88N/+ retinas, positive RHO staining was observed within the whorls and rosettes; in CrxK88N/N mutant retinas, RHO staining was completely absent. Importantly, we did not observe mis-localized RHO staining in the inner retinal layers suggesting that the developmental programs of other retinal cell types were not directly affected by E80A or K88N mutation. Overall, abnormalities in the cone/rod gene expression matched the corresponding human disease diagnosis20,22, and the phenotypic severity correlated with the degree of mis-regulation of CRX target genes in the corresponding RNAseq dataset. Thus, these results support a model that CRX HD mutation-mediated mis-regulation of gene expression disrupts photoreceptor terminal differentiation and leads to defects in retinal layer organization and OS formation.

E80A and K88N mouse models show visual function deficits that recapitulate human diseases

To understand the consequences of disrupted photoreceptor differentiation on visual function, we measured electroretinogram (ERG) responses to light stimuli for WT and mutant mice at one month of age (Figures 5K-M). In response to incremental changes of light intensities, WT animals showed corresponding amplitude increases in dark adapted A-waves (rod signals) and B-waves (rod-evoked bipolar cell signals), as well as in light-adapted B-waves (cone-evoked bipolar cell signals). The three severe mutants, CrxE80A/A, CrxK88N/+, and CrxK88N/N had no detectable dark-adapted or light-adapted ERG responses, suggesting that these mice have no rod or cone function and are blind at young ages. The null ERG phenotype of the CrxK88N animals is consistent with the clinical LCA phenotype in humans22. In contrast, CrxE80A/+ animals retained partial rod ERG responses as indicated by the reduced A-wave and B-wave amplitudes (Figures 5K and 5L). Yet, CrxE80A/+ animals had no detectable cone ERG responses, which is consistent with the CoRD clinical phenotype in humans20 (Figure 5M). Taken together, the visual function impairment in each CRX HD mutant model, coincided with the morphological and molecular changes, suggesting that CrxE80A and CrxK88N mouse models recapitulate the corresponding human diseases.

CRX E80A has increased transactivation activity and leads to precocious differentiation in CrxE80A retinas

Lastly, we asked what might be the molecular mechanism that causes the mis-regulation of photoreceptor genes in the mutant retinas. Previous studies have established that reporter assays with the rhodopsin promoter in HEK293T cells can measure changes of CRX transactivation activity and inform the mechanisms by which photoreceptor genes are mis-regulated in CRX mutant retinas17. We thus tested the transactivation activity of the three CRX HD mutants on the pRho-Luc reporter in HEK293T cells (Figure 6A, STAR Methods). Consistent with published studies, R90W mutant had significantly reduced activity compared to WT CRX17. K88N mutant showed a similarly reduced activity as R90W, consistent with K88N’s loss of binding at canonical CRX sites and failure to activate CRX-DAGs in vivo (Figures 2A, 2B, 3A-3C). In contrast, E80A mutant, which binds to canonical CRX sties, showed significantly increased transactivation activity on the Rho promoter. This hyperactivity of E80A protein at Rho promoter correlates with the upregulation of CRX-DAGs in the mutant retinas at P10 (Figures 3A-3C).

CRX E80A hyperactivity underlies precocious photoreceptor differentiation in CrxE80A retinas.

(A) Boxplot showing luciferase reporter activities of different CRX variants. P-values for one-way ANOVA with Turkey honestly significant difference (HSD) test are indicated. p-value: ****: ≤0.0001, ***: ≤0.001, ns: >0.05. (B-D) Rhodopsin (RHO, green) immunostaining is absent in P3 WT retina but detected in CrxE80A/+ and CrxE80A/A retinas. Nuclei are visualized by DAPI staining (Blue). Arrow indicates the sporadic RHO staining in CrxE80A/+ sample. ONBL: outer neuroblast layer; GCL: ganglion cell layer. Scale bar, 100μm.

To determine the consequences of E80A hyperactivity on photoreceptor differentiation timing, we compared RHO protein expression during early postnatal retinal development using retinal section immunostaining (Figures 6B-6D). In WT retinas, most rods were born by P3 but had not differentiated (Figure 6B). Previously studies showed that RHO proteins were detected by IHC starting around P7 in WT retinas63. In comparison, both CrxE80A/+ and CrxE80A/A retinas showed positive RHO staining at P3 (Figures 6C and 6D). RHO+ cells were largely seen in the outer portion of the ONBL layer in CrxE80A/+ retinas, and strikingly spread throughout the large presumptive ONL layer in CrxE80A/A retinas. The detection of RHO protein in P3 CrxE80A mutant retinas indicates that photoreceptor differentiation program was precociously activated. Taken together, our results support a model that E80A and K88N mutations each perturbs CRX regulatory activity in a unique way, causes photoreceptor differentiation defects, and ultimately leads to distinct dominant disease phenotypes that recapitulate human diseases.

Discussion

Through molecular characterization of mutant proteins, transcriptome and cellular profiling of developing mutant mouse retinas and ERG testing of adult retinas, we have identified two novel pathogenic mechanisms of CRX HD mutations, E80A and K88N, that are associated with dominant CoRD and dominant LCA in human20,21. Distinct from the previously characterized loss-of-function R90W mutation18,19, E80A and K88N mutations produce altered CRX proteins with gain of regulatory functions -CRX E80A is associated with increased transcriptional activity and CRX K88N has altered DNA binding specificity. Both CRX E80A and CRX K88N proteins impair photoreceptor gene expression, development and produce structural and functional deficits in knock-in mouse models, recapitulating human diseases64 (Figure 7). Thus, precise regulatory activity and target specificity of CRX are both essential for proper photoreceptor development and functional maturation.

Missense mutations in CRX HD affect photoreceptor gene expression and leads to distinct retinal disease phenotypes through gain- and loss-of-function mechanisms.

The CrxE80A and CrxK88N knock-in mouse models recapitulate dominant human CRX-linked retinopathies and present novel morphological deficits. Published Crx mutant mouse models have established that decreased expression of CRX or loss of CRX transcriptional activity results in incomplete photoreceptor differentiation and ultimately rapid degeneration of immature photoreceptors15,18. Despite the difference in molecular mechanisms, both CrxE80A and CrxK88N mouse models show defective photoreceptor terminal differentiation and develop whorls and rosettes in the ONL layers by P21, which are not observed in degenerative CRX mouse models18. No significant increase in cell death is observed in CrxE80A and CrxK88N mutant mouse retinas at P21, suggesting distinct pathogenic mechanisms than the photoreceptor degeneration previously reported in other CRX mouse models. P10 RNAseq analysis on cell markers of other neuronal cell types and Muller glia cells did not detect significant changes, suggesting that the developmental programs for these cell types were unaffected. Further studies are needed to systematically characterize changes in photoreceptor differentiation and retinal morphology during early development, and interactions between diseased photoreceptors and other major cell types in CrxE80A and CrxK88N mutant mouse retinas.

Precisely tuned CRX-mediated transcriptional activity is important to ensure proper photoreceptor development. Previous studies have demonstrated that CRX is activated shortly after cell cycle exit in retinal progenitor cells fated to be photoreceptors66,67. CRX then works with other cell type specific TFs, such as NRL and NR2E3 (rods) or RORB and THRB (cones), to regulate gene expression important for cell-type specific structures and functions and drive the differentiation of photoreceptor precursors68,69. Loss-of-function animal studies have established that CRX is primarily a transcriptional activator since loss of CRX protein or regulatory function is associated with decreased expression of photoreceptor-specific genes regulated by CRX19. Here, we find that CRX E80A mutant, that displays increased CRX transcriptional activator activity, is also deleterious to photoreceptor development. CRX E80A maintains WT DNA-binding specificity in vitro and binds to canonical CRX sites in mutation knock-in mouse retinas. Consistent with CRX E80A’s hyper-activity at canonical CRX target pRho, in differentiating CrxE80A/+ and CrxE80A/A retinas, many of the CRX-dependent activated genes implicated in rod functions are prematurely up-regulated at early postnatal ages. The RHO protein is also detected in developing rod photoreceptor in CrxE80A/+ and CrxE80A/A retinas as early as P3. However, at later ages, rod photoreceptor terminal differentiation is impaired, resulting in disrupted retinal morphology and defective visual functions. Photoreceptor differentiation is precisely programmed via sequential and concerted gene expression programs within a defined time window66,67. One explanation for adult CrxE80A phenotypes is that CRX E80A associated hyper-activity precociously activates later stage genes in the absence of proper nuclear context and/or subcellular structures, which in turn negatively impacts early events in photoreceptor differentiation. Multiple possibilities could account for CRX E80A associated hyper-activity: (1) E80A mutation increases CRX interactions with its binding sites; (2) E80A mutation increases CRX interactions with the transcription complex or other TFs; (3) certain protein factors that could promote photoreceptor gene expression are up-regulated in CRX E80A expressing cells.

While these possibilities are not mutually exclusive, evidence from the current study favors the first possibility since E80A HD is less discriminative of HD motif variants than WT HD as shown by Spec-seq (Figure 1I). Regulatory sequences of many photoreceptor genes contain both consensus and non-consensus CRX motifs13,14,17,34. CRX E80A might interact better with non-consensus CRX motifs within binding sites, resulting in increased transcriptional output of target genes. Large scale reporter assays (e.g. Massively Parallel Reporter Assays) in the appropriate cellular context – neonatal mouse retinas – would be informative to understanding CRX E80A hyper-activity mechanism and the direct impacts of its hyper-activity.

Although a global increase in expression is expected in CrxE80A models, a subset of CRX dependent activated genes implicated in cone photoreceptor development and functions is down-regulated in both differentiating and mature mutant retinas. Cones are born prenatally in mice within an earlier time window than rods. Yet, cones undergo terminal differentiation in a similar postnatal window as rods to develop into subtypes of cones – M-cones v.s. S-cones. One model is that CRX E80A improperly activates later stage cone genes at a much earlier time window, disrupting cone terminal differentiation. Supporting this model, similar to the presented pRho data, CRX E80A also hyper-activates the S-cone opsin promoter (data not shown).

Detecting premature activation of cone genes in CrxE80A retinas will require analysis at early developmental ages. An alternative model for lack of cone markers in CrxE80A retinas is that cones might be more sensitive to perturbations in CRX transcriptional activity, which impacts cone survival. It is known that cones depend on a different repertoire of TFs than rods for subtype terminal differentiation59. It is possible that cone TFs respond differently to mutant CRX E80A hyper-activity, leading to the distinct expression changes in CrxE80A mutant retains. Since cones only make up a very small portion of photoreceptors in mouse retinas70, characterization of CRX E80A activity in purified cone population at earlier ages will be necessary to understand its selective effect on the cone differentiation program and help to elucidate WT CRX regulatory principles in early photoreceptor development. Collectively, CRX transcriptional activity at its canonical genomic binding sites need to be maintained within a quantitatively defined range such that CRX dependent target genes are activated in a precise temporal pattern to ensure proper photoreceptor differentiation.

CRX DNA binding specificity constitutes an additional regulatory program required for the fidelity of photoreceptor gene expression during development. Previous work has shown that subtle differences in HD motif content could have distinct transcriptional and functional outcomes. For instance, studies in D. melanogaster demonstrate that the sequential use of HD DNA motif variants defines different classes of enhancers and coordinates the precise timing of gene expression patterns during embryonic development71. In our current study, the K88N mutation alters CRX DNA-binding specificity that redirects CRX binding to binding sites with N88 HD motifs in the mutant mouse retinas. CrxK88N retinas show more severe perturbations in photoreceptor gene expression and development than in loss-of-function mutant CrxR90W/W. Concomitantly, CrxK88N/+ and CrxK88N/N mice display the most severe photoreceptor morphological deficits observed in any Crx mouse models and show absence of visual functions in young adults. In developing WT mouse retinas, the HD motif preferred by HD TFs with a glutamine (Q) at HD50 position encodes quantitatively different activity than the CRX consensus motif suggesting functional difference between HD binding site variants60,72. It is likely that the severe CrxK88N phenotypes are attributed to both loss of activity at canonical CRX motifs and ectopic binding and transcriptional activity of CRX K88N proteins at N88 HD motifs. Additionally, since a functional WT copy of CRX is retained in CrxK88N/+ retinas and CRX K88N prefers different motifs than WT CRX, the lack of WT activity alone cannot explain the severe expression and developmental deficits.

Alternatively, these results suggest involvement of additional regulatory mechanisms: CRX K88N activity at N88 HD motifs might (1) ectopically activate genes whose expression prevents the progression of development or inactivate genes required for development; (2) interfere with other HD TFs that also recognize N88 HD motifs; (3) lead to epigenetic alterations that antagonize normal CRX functions. Many other HD containing TFs are expressed in developing mouse retina, including OTX2, RAX, VSX2, PAX6, SIX3/6, and LHX family6. Different from CRX, these HD TFs are essential for gene regulation in retinal progenitor cells and/or in other retinal cell lineages. It is very likely that CRX correctly targets photoreceptor genes at least through its DNA binding specificity and alteration of CRX DNA binding specificity would mis-regulate genes originally targeted by other HD TFs and lead to perturbations in GRNs and development. To date, most studies have focused on CRX activity at cis-regulatory sequences enriched for the WT CRX consensus motifs73,74. Systematic comparison of transcriptional activity at N88 HD motifs and WT consensus motifs in the context of photoreceptor development in both WT and mutant retinas would be needed to substantiate the impact of mutant CRX K88N activity at different HD motifs. These experiments will also help clarify the pathogenic mechanisms in the CrxK88N models and extend our knowledge of CRX regulatory grammar and activation of photoreceptor genes during development.

One limitation of this work is that effects of E80A and K88N mutations on CRX HD-DNA interactions have been evaluated at monomeric HD motifs and with homogenous protein species both in vitro and in vivo. Further evaluation of WT and mutant CRX binding at dimeric motifs will be desirable, since selected dimeric HD motifs are known to mediate HD TF cooperativity to ensure gene expression fidelity during development75,76. Relatedly, we also need to address how CRX WT and mutant E80A or K88N proteins interact at HD binding motifs – whether they cooperate or compete with each other, whether these interactions are HD motif sequence-dependent, and how gene expression is impacted by CRX cooperativity or competition. While CRX HD mediates both TF-DNA interactions and protein-protein interactions, evaluation of how E80A and K88N mutations impact CRX interaction with other important photoreceptor TFs and how perturbations in these interactions lead to disease phenotypes warrant further study.

Since the spatial structures and HD-DNA contact models of HD proteins are evolutionarily conserved, our study of CRX provides valuable molecular insights for HD mutations implicated in other diseases. For example, p.E79K substitution (corresponds to CRX E80) in OTX2 HD is associated with dominant early-onset retinal dystrophy in two Caucasian Canadian families77, heterozygous p.R89G (corresponds to CRX R90) mutation in OTX2 HD causes severe ocular malformations78, and missense mutations of the CRX K88 and R90 homologous residues in PITX2 HD are associated with Rieger syndrome79. It is likely that these mutations affect HD activity in similar ways as observed in CRX, and the exact disease manifestation is determined by cell-type specific mechanisms. The retina is readily accessible and a broad range of molecular tools are available for ex vivo and in vivo manipulations. We believe that CRX is an ideal model to understand the pathogenic mechanisms of HD mutations and to test therapeutic regimens, which would ultimately benefit the study of HD TFs and their associated diseases in other tissues and organs.

Collectively, our findings support a unifying model in which precise CRX interaction with cis-regulatory sequences is essential for gene expression and functional maturation during photoreceptor development. Mutations that perturb CRX target specificity or transactivation activity impair the integrity of CRX-mediated transcriptional regulation and result in distinct morphological and functional deficits (Figures 7A-7D). Regulatory activity at ectopic N88 HD motifs antagonizes normal CRX functions and leads to aberrant photoreceptor development and early vision loss. This model also emphasizes the importance of tailoring gene therapy regimens to tackle individual pathogenic mechanisms. For instance, while supplementing WT CRX might be sufficient to rescue a hypomorphic/loss-of-function mutant, simultaneous elimination of a gain-of-function CRX product would be necessary to rescue dominant mutants, as exemplified in a recent report of allele-specific gene editing to rescue dominant CRX-associated LCA7 phenotypes in a retinal organoid model80. We believe that this principle also applies to other dominant neurological diseases. Additionally, with the refinement of the CRX mechanistic model, when new disease mutations are identified, genetic counsellors can now provide more informed predictions of disease progression and future visual deficits. This information is important for individuals to be psychologically prepared and seek necessary assistance to improve their quality of life.

eLife Assessment

This manuscript is of interest to readers in the field of neural development. The study is important as it examines two mutations within a homeodomain transcription factor called Cone-Rod Homeobox (CRX) that cause retinopathy in humans. The data are solid, and the work contributes to our understanding of the underlying pathogenetic mechanisms.

Reviewer #1 (Public Review):

The manuscript by Zheng et al. examined the disease-causing mechanisms of two missense mutations within the homeodomain (HD) of CRX protein. Both mutations were found in humans and can produce severe dominant retinopathy. The authors investigated the two CRX HD mutants via in vitro DNA-binding assay (Spec-seq), in vivo chromatin-binding assay (ChIP-seq), in vivo expression assay of downstream target genes (RNA-seq), and retinal histological and functional assays. They concluded that p.E80A increased the transactivation activity of CRX and resulted in precocious photoreceptor differentiation, whereas p.K88N significantly changed the binding specificity of CRX and led to defects in photoreceptor differentiation and maintenance. The authors performed a significant amount of analyses. The claims are sufficiently supported by the data. The results not only uncovered the underlying disease-causing mechanisms, but also can significantly improve our understanding of the interaction between HD-TF and DNA during development.

Minor concerns:

1. The E80A, K88N and R90W (previously reported by the same group) mutations are located very close to each other in the homeodomain (Figure 1A), but had distinct effects on the activity of CRX. Has the structure of the homeodomain (of CRX) been resolved? If so, could the authors discuss this phenomenon (mutations close to each other but have distinct effects) based on the HD-DNA structure? In addition, has this phenomenon been observed in other homeodomain TFs?

2. The authors should briefly summarize the effects/disease-causing-mechanisms of all the reported CRX mutations in the discussion part. The readers can then have a better overview of the topic.

3. CRX can also function as a pioneer factor (reported by the same group). Would these HD mutations distinctively affect chromatin accessibility (which then leads to ectopic binding on the genome)?

4. The discussion part can be shortened and simplified.

Reviewer #2 (Public Review):

Zheng et al., investigated the molecular and functional mechanisms of two homeodomain missense mutations causing human retinal photoreceptor degeneration diseases in photoreceptor development regulated by the CRX transcription factor. They analyzed the E80A mutation associated with dominant cone-rod dystrophy (CRD) and the K88N mutation associated with dominant Leber Congenital Amaurosis (LCA). The authors found that E80A CRX binds to the same target DNA sites as WT CRX, but the binding specificity of K88N CRX is altered from that of WT in an in vitro assay. They generated Crx(E80A) and Crx(K88N) KI mice and performed ChIP assay and observed that K88N CRX binds to novel genomic regions from the WT-binding sites, while E80A binds to the WT sites. In addition, using the KI mice, they found that E80A and K88N differently affect the expression of Crx target genes. This study is well executed with proper and solid methodologies, and the manuscript is clearly written. This study gives us the insights how single missense CRX mutations lead to different types of human retinal photoreceptor degeneration diseases.

While the study has strengths in principle, it has a couple of weaknesses. One is how well E80A KI mice function as a pathological model of dominant CRD, in which cones are mainly first affected, is not clearly shown in this study. More data investigating how cones are affected by performing histological, molecular, and physiological analyses will be helpful and useful. For example, in the Discussion, the authors describe that E80A associates with S-cone opsin promoter results is "data now shown". This data must be presented for the readers. In addition, more molecular insights as to how E80A affects cones will strengthen this study. Another point is that it will be very valuable if the authors could show how E80A and K88N differently affect the 3D structure of the CRX homeodomain. Even a simulation model would be valuable.

Author Response:

The following is the authors' response to the original reviews.

Thank you for sending our manuscript for review and the positive editorial comments. On behalf of all authors, I would like to thank the reviewers for their critical reading of our manuscript and for providing insightful and valuable suggestions. We have revised the discussion section accordingly, including a new supplemental figure to show the results previously stated as “data not shown”. Please see below for detailed explanations.

Reviewer #1 (Public Review):

The manuscript by Zheng et al. examined the disease-causing mechanisms of two missense mutations within the homeodomain (HD) of CRX protein. Both mutations were found in humans and can produce severe dominant retinopathy. The authors investigated the two CRX HD mutants via in vitro DNA-binding assay (Spec-seq), in vivo chromatin-binding assay (ChIP-seq), in vivo expression assay of downstream target genes (RNA-seq), and retinal histological and functional assays. They concluded that p.E80A increased the transactivation activity of CRX and resulted in precocious photoreceptor differentiation, whereas p.K88N significantly changed the binding specificity of CRX and led to defects in photoreceptor differentiation and maintenance. The authors performed a significant amount of analyses. The claims are sufficiently supported by the data. The results not only uncovered the underlying disease-causing mechanisms, but also can significantly improve our understanding of the interaction between HD-TF and DNA during development.

Thank you for summarizing the key findings and strengths of our manuscript.

Minor concerns:

1. The E80A, K88N and R90W (previously reported by the same group) mutations are located very close to each other in the homeodomain (Figure 1A), but had distinct effects on the activity of CRX. Has the structure of the homeodomain (of CRX) been resolved? If so, could the authors discuss this phenomenon (mutations close to each other but have distinct effects) based on the HD-DNA structure?

In paragraphs 2, 4, 5 of the discussion section, we have added explanations on how each mutation could affect CRX HD-DNA interactions differently based on published structural studies. And we further explain how these biochemical changes relate to the molecular perturbations and cellular phenotypes seen in vivo.

In addition, has this phenomenon been observed in other homeodomain TFs?

Disease associated missense mutations at residues HD50 (K88) and HD52 (R90) have also been reported in other HD TFs implicated in CNS development (see discussion paragraph 7). Distinctively, different substitutions at CRX E80 residue have been reported in multiple CoRD cases, suggesting its essential role in HD-DNA-mediated regulation during retinal development. These new points are now included in the discussion section.

2. The authors should briefly summarize the effects/disease-causing-mechanisms of all the reported CRX mutations in the discussion part. The readers can then have a better overview of the topic.

We have added a concise summary of previously proposed CRX mutation classification scheme, all characterized Crx mutant mouse models and their pathogenic mechanisms. Please see paragraph 9 in the discussion section.

3. CRX can also function as a pioneer factor (reported by the same group). Would these HD mutations distinctively affect chromatin accessibility (which then leads to ectopic binding on the genome)?

Prior evidence has demonstrated that regulatory regions for many photoreceptor genes failed to stay accessible upon loss of CRX in the Crx-/- model (PMID: 30068366). It is unclear with the existing data whether CRX could initiate the chromatin remodeling (true pioneering function) of these regions, or it simply maintains the accessibility once these regions became accessible. Future studies comparing epigenomic landscape changes in mutant Crx KI models at various ages can be informative, particularly for the CRX K88N ectopic binding events. Determining how the CRX K88N mutant protein alters chromatin landscape important for photoreceptor fate and/or differentiation during development would shed light on the nature of these ectopic binding events.

4. The discussion part can be shortened and simplified.

We have re-written the discussion section to make it concise and to incorporate discussions on mutant CRX HD structures. Please see the revised manuscript.

Reviewer #2 (Public Review):

Zheng et al., investigated the molecular and functional mechanisms of two homeodomain missense mutations causing human retinal photoreceptor degeneration diseases in photoreceptor development regulated by the CRX transcription factor. They analyzed the E80A mutation associated with dominant cone-rod dystrophy (CRD) and the K88N mutation associated with dominant Leber Congenital Amaurosis (LCA). The authors found that E80A CRX binds to the same target DNA sites as WT CRX, but the binding specificity of K88N CRX is altered from that of WT in an in vitro assay. They generated Crx(E80A) and Crx(K88N) KI mice and performed ChIP assay and observed that K88N CRX binds to novel genomic regions from the WT-binding sites, while E80A binds to the WT sites. In addition, using the KI mice, they found that E80A and K88N differently affect the expression of Crx target genes. This study is well executed with proper and solid methodologies, and the manuscript is clearly written. This study gives us the insights how single missense CRX mutations lead to different types of human retinal photoreceptor degeneration diseases.

We greatly appreciate the reviewer’s summary and positive comments.

While the study has strengths in principle, it has a couple of weaknesses. One is how well E80A KI mice function as a pathological model of dominant CRD, in which cones are mainly first affected, is not clearly shown in this study. More data investigating how cones are affected by performing histological, molecular, and physiological analyses will be helpful and useful. For example, in the Discussion, the authors describe that E80A associates with S-cone opsin promoter results is "data now shown". This data must be presented for the readers. In addition, more molecular insights as to how E80A affects cones will strengthen this study.

The mouse retina is rod dominant and contains only a small number of cones (3% of all photoreceptors) that are born prenatally. This poses technical challenges to appropriately assess cone- specific changes during disease initiation/progression. We are in the process of developing cellular/molecular tools to investigate how cones are being affected in Crx E80A KI model, but this is beyond the scope of the current study.

At the same time, we have added a supplemental panel showing that, based on P0 retinal immunostaining of the early cone marker RXRγ, cones were initially born, and fate specified in CrxE80A retinas (see Figure S7A). Since the E80A protein also hyper-activated S-cone opsin promoter-luciferase (Sop-luc) reporter in HEK293 cells (see Figure S7B), we predict that CRX E80A affects cone photoreceptor differentiation in a similar manner as rod photoreceptors. Furthermore, the cone transcriptional program might be more prone to perturbations by abnormal CRX activities. These possibilities require future investigations. For this manuscript, we have included all these points in the discussion section.

Another point is that it will be very valuable if the authors could show how E80A and K88N differently affect the 3D structure of the CRX homeodomain. Even a simulation model would be valuable.

Please see our answer to Point 1 of Reviewer #1. In short, we have added in the discussion section our explanations on how each mutation could affect CRX HD-DNA interactions differently based on structural studies. We further explain how these biochemical changes relate to the molecular perturbations and cellular phenotypes seen in vivo. Additionally, since TF-DNA interactions are diverse and dynamic across binding sites with different sequence features and genomic environments, future studies that systematically and quantitatively evaluate CRX transcriptional activity at different regulatory sequences would be important.

Recommendations for the authors:

Reviewer #2 (Recommendations For The Authors):

As a minor comment, in page 8, second section, "Previous studies have demonstrated the CRX is activated shortly after cell cycle exit in retinal progenitor cells fated to be photoreceptor.", the authors cited refs 66 and 67, which were in 2105 and 2016. However, this was demonstrated in the paper of J. Neurosci.31(46), 16792-807, 2011, Figure 1. It would be fair for the authors to cite the JN 2011 paper.

Thanks to the reviewer for the suggested reference, we have added it to the revised manuscript.

Version 2

Abstract

Summary

Homeodomain transcription factors (HD TFs) are instrumental to vertebrate development. Mutations in HD TFs have been linked to human diseases, but their pathogenic mechanisms remain elusive. Here we use Cone-Rod Homeobox (CRX) as a model to decipher the disease-causing mechanisms of two HD mutations, p.E80A and p.K88N, that produce severe dominant retinopathies. Through integrated analysis of molecular and functional evidence in vitro and in knock-in mouse models, we uncover two novel gain-of-function mechanisms: p.E80A increases CRX-mediated transactivation of canonical CRX target genes in developing photoreceptors; p.K88N alters CRX DNA-binding specificity resulting in binding at ectopic sites and severe perturbation of CRX target gene expression. Both mechanisms produce novel retinal morphological defects and hinder photoreceptor maturation distinct from loss-of-function models. This study reveals the distinct roles of E80 and K88 residues in CRX HD regulatory functions and emphasizes the importance of transcriptional precision in normal development.

Introduction

Homeodomain transcription factors (HD TFs) play a fundamental role in vertebrate development. Members of the HD TF family are characterized by the presence of a highly conserved 60 amino acid helix-turn-helix DNA-binding domain known as the homeodomain (HD). The HD is one of the most studied eukaryotic DNA-binding motifs since its discovery in Drosophila homeotic transformations1, 2. Hundreds of HD TFs have subsequently been documented in regulating gene expression programs important for body plan specification, pattern formation and cell fate determination1, 2. Mutations in HD TFs have been linked to many human diseases, including neuropsychiatric and neurodegenerative conditions3, 4. Although significant progress has been made in understanding HD-DNA interactions, uncovering the pathogenetic mechanisms of disease-causing missense mutations in HD have proven challenging.

The retina has long been used as a model system to study the role of HD TFs in normal central nervous system (CNS) development and in neurological diseases5. During retinogenesis, HD TFs play essential roles in the patterning of neuroepithelium, specification of retinal progenitors and differentiation of all retinal cell classes that derive from a common progenitor6. Importantly, many HD TFs are shared between the brain and the retina during development and mutations in these TFs can lead to disease manifestation in both tissues7–12. The accessibility and wealth of available molecular tools makes the retina a valuable tool to decipher the pathogenic mechanisms of HD TF mutations associated with neurological diseases.

Here, we study CRX, a HD TF essential for photoreceptor cells in the retina, as a model to understand how single amino acid substitutions in the HD impact TF functions and cause blinding diseases. Photoreceptors are the most numerous neurons in the retina and are specialized to sense light and initiate vision through a process called phototransduction. Animal studies have demonstrated that Crx is first expressed in post-mitotic photoreceptor precursors and maintained throughout life13–15. Loss of CRX results in impaired photoreceptor gene expression, failure of maturation and rapid degeneration of immature, non-functional photoreceptors16. Protein-coding sequence variants in human CRX have been associated with inherited retinal diseases (IRDs) that affect photoreceptors: Leber congenital amaurosis (LCA), cone-rod dystrophy (CoRD), and retinitis pigmentosa (RP) (OMIM:602225). However, the recessive phenotype observed in Crx knockout mouse models fails to recapitulate many dominant human CRX mutations that arise de novo16.

CRX contains two functional domains – the N-terminal HD and C-terminal activation domain (AD) (Figure 1A), both are required for proper activation of target genes and maintenance of normal Crx mRNA transcript abundance17, 18. To understand how CRX HD mutations cause diseases, we have previously reported a mutation knock-in mouse model carrying a hypomorphic mutation p.R90W (R90W) in CRX HD19, 20. We found that R90W mutation produces a recessive loss-of-function phenotype very similar to that of Crx knockout mice.

Disease associated missense mutations altered CRX HD DNA binding specificity.

(A) Diagram of CRX functional domains: Homeodomain (HD) for DNA-binding and Activation Domain (AD) for target gene transactivation. The three missense mutations in this study are located at the C-terminus of CRX HD and associated with different retinal diseases in human. Number in the parenthesis denotes the CRX HD position of the corresponding mutated residue. (B) Alignments of HD recognition helix sequences for the indicated HD proteins for which HD missense mutations have been associated with inherited diseases. Accession numbers can be found in Supplementary Table S1. Missense variants in this study (highlighted) are located at highly conserved residues across species and different HD TFs. (C) Spec-seq experimental workflow (Methods). (D) Spec-seq library design of monomeric HD binding sites. (E) EMSA gel images of Spec-seq experiments with different CRX HD species. Bx: Bound. B-: Unbound. (F) Relative binding energy comparison from two different experiments with WT HD. (G) Binding energy model for WT CRX HD. (H-J) Relative binding energy comparison between WT HD and R90W HD (H), E80A HD (I), or K88N HD (J). Consensus sequence is defined to have relative binding energy of 0kT (TAATCC for WT, R90W and E80A, TAATTA for K88N). The identity line is represented in grey dash. The orange dashed line shows the best linear fit to the data. (K-M) Binding energy models for R90W HD (K), E80A HD (L), and K88N HD (M). Only sequence variants within two mismatches to the corresponding consensus sequences were used to generate binding models. Negative binding energy is plotted such that bases above the x-axis are preferred bases and bases below the x-axis are unfavorable bases. Constant bases (TAA) carried no information are drawn at arbitrary height in grey.

Intriguingly, several missense mutations within the same HD recognition helix as R90W, including p.E80A (E80A) and p.K88N (K88N), are linked to severe dominant IRDs21–23 (Figures 1A and 1B). Here, we utilized a multi-omics approach to investigate the functional consequences of the E80A and K88N mutations on CRX regulatory activities and photoreceptor development (Figure S1). Comparison of the in vitro HD-DNA binding models of CRX and disease variants generated by Spec-seq revealed unique specificity changes of each mutant protein. Introduction of each mutation into the endogenous Crx locus generated knock-in mouse models CrxE80A and CrxK88N that reproduced dCoRD- and dLCA-like phenotypes. ChIP-seq analysis of CRX binding in vivo revealed mutation-specific changes in CRX targetome, consistent with mutation-specific DNA binding changes in vitro. Retinal RNA-seq analysis uncovered two distinct mechanisms by which the two HD missense mutations contribute to altered gene expression programs during photoreceptor differentiation and maturation. Our results highlight the importance of residues E80 and K88 in CRX-mediated transcriptional regulation during photoreceptor development and the diverse mechanisms by which HD missense mutations can affect TF functions and lead to severe dominant neurological diseases.

Results

K88N but not E80A mutation alters CRX HD DNA-binding specificity in vitro

CRX belongs to the paired-like HD TF family that recognize a 6-bp DNA motif in a stereotypic way24–30. Extensive studies of the HD have revealed a canonical HD-DNA recognition model where recognition of the 3’ region (bases 4-6) of the HD DNA binding site is mediated by specificity determinants within the conserved HD recognition helix, corresponding to CRX residues 80-9624–27, 30, 31 (Figure 1B). In particular, HD residue 50, equivalent to CRX K88 residue (Figure 1A), is the major specificity determinant in paired-like HD TF-DNA interactions25, 26. Since the three disease-associated HD missense mutations, E80A, K88N and R90W, are located within the CRX HD recognition helix, we wondered if these mutations change CRX HD DNA-binding specificity.

We adapted a high-throughput in vitro assay, Spec-seq, that determines protein-DNA-binding specificity by sequencing32–34. Spec-seq was developed based on the traditional electrophoretic mobility shift assay (EMSA) to measure protein-DNA interactions. Spec-seq allows us to measure the relative binding affinities (i.e., specificity) for a library of HD binding motifs in parallel and generate quantitative binding models for different CRX HDs (Figures 1C). Based on the HD-DNA interaction model, we designed and tested a Spec-seq library containing all possible monomeric HD motifs (TAANNN) (Figure 1D).

We first obtained the wild-type (WT) CRX HD DNA binding model with Spec-seq using bacterially-expressed and affinity-purified HD peptides (Figures 1E-1G, Methods). Relative binding energies of CRX WT HD from two experiments showed strong correlation (r: 0.984) and noise level (0.114 kT) within the expected range in typical Spec-seq data (Figure 1F). Binding energy model of WT HD was then generated by applying multiple linear regressions on the relative binding energies of all sequences within two base-pair mismatches to the WT CRX consensus (TAATCC)13 (Figure 1G, Methods). A clear preference for CC bases at the 3’ end of the motif is consistent with known CRX binding preference in vitro and in vivo35, 36.

We next sought to understand how disease mutations affect CRX DNA-binding specificity. We purified all mutant HD peptides following the same protocol as WT HD peptides and verified their DNA binding (Figures S2A-S2D)18. Comparison of the relative binding energies between each pair of mutant and WT HD revealed distinct effects (Figures 1H-1J). By definition, the consensus DNA binding motif of a testing peptide has a relative binding energy of 0kT. The relative binding energy difference between nucleotide variants and the consensus motif correlates with the DNA binding specificity of the testing peptide. We found that R90W HD and E80A HD both prefer the same consensus motif as WT (Figures 1K and 1L). R90W HD bound with slightly higher specificity than WT, as demonstrated by most data points falling above the identity line (Figures 1H and 1K), suggesting that R90W HD is more sensitive than WT to binding sequence variations. In contrast, when comparing E80A HD with WT HD, many data points fell below the identity line and the relative binding energies regressed towards 0 on the E80A axis (Figures 1I and 1L). This suggests that E80A HD bound with lower specificity than WT HD and thus was more tolerant to base variations in the HD DNA motif. Different from R90W and E80A, K88N mutation dramatically altered CRX HD DNA-binding specificity (r: 0.160) (Figures 1J and 1M). The K88N preferred binding sequence (TAAT/ATT/A) is referred to as N88 HD motif hereafter. K88N HD also had the largest degree of discrimination from its preferred to the weakest binding motif, suggesting that it is most sensitive to variants in the HD DNA motif. As a control, we tested a second library with the TAANNN sites on the reverse strand and obtained similar results (Figures S2E-S2L). Together, these results indicate that while E80A mutation does not affect CRX HD DNA binding specificity, the K88N mutation dramatically alters the specificity in vitro.

E80A protein binds to WT sites while K88N occupies novel genomic regions with N88 HD motifs in vivo

Next, we asked if changes in DNA-binding specificity affected mutant CRX chromatin binding in developing photoreceptors. We first created two human mutation knock-in mouse models, CrxE80A and CrxK88N, each carrying a single base substitution at the endogenous Crx locus (Figures S3A and S3B, Methods). We confirmed that Crx mRNA was expressed at comparable levels in WT and mutant retinas (Figure S3C), and the full-length CRX proteins were readily detectable in the nuclear extracts from all samples (Figure S3D). We then obtained genome-wide binding profiles for each CRX variant by chromatin immunoprecipitation followed by sequencing (ChIP-seq) on mouse retinas at P14, a time when all retinal cell types are born, photoreceptor specification is completed in WT animals, and prior to any observed cell death in other CRX mutants previously characterized19, 37. To focus on changes specific to each mutant CRX protein, only homozygous animals were used for ChIP-seq profiling.

Unsupervised clustering of all CRX binding sites revealed two major clusters (Figure 2A, Methods). Cluster 1 consisted of canonical WT CRX binding sites that are also occupied by CRX E80A protein (Figures 2A and 2B). Similar to WT CRX, CRX E80A binding in vivo was mostly enriched in intronic, followed by intergenic and TSS regions (Figure 3D). In contrast, CRX R90W, that also showed similar consensus preference to WT in vitro, failed to produce significant DNA binding in vivo (Figures 2A-2C, Methods). This suggests that the retinopathy phenotype of CrxR90W/W is likely due to loss of binding at canonical WT CRX binding sites. Intriguingly, while CRX K88N showed loss of binding at canonical CRX binding sites, it gained a small set of binding sites (Cluster 2, Figures 2A-2C). De novo motif searching with DREME38 under CRX peaks in each genotype revealed enrichment of monomeric HD motifs (Figure 2E) consistent with those found in Spec-seq (Figures 1G and 1K-1M), highlighted by a change in enriched HD motif from WT CRX HD type to N88 HD type in the CrxK88N/N retinas. Consistency with in vitro binding models suggests that in vivo changes in CRX chromatin binding were at least in part driven by the intrinsic changes in HD-DNA binding specificity by each individual mutation.

CRX E80A binds to WT sites while CRX K88N occupies novel genomic regions enriched for N88 HD motif in vivo.

(A) Enrichment heat map depicting CRX ChIP-seq normalized reads centered at all possible CRX peaks ± 2kb, sorted by binding intensity in WT samples. Clusters were defined by hierarchical clustering of CRX binding intensity matrix from all genotypes (STAR Methods). (B-C) Genome browser representations of ChIP-seq normalized reads for different CRX species in P14 WT and mutant mouse retinas at Rho and Atf2. (D) Enrichment heatmap showing fraction of CRX ChIP-seq peaks fall in different genomic environments. (E) Logo representations of de novo found short HD motifs under CRX ChIP-seq peaks in WT and mutant mouse retinas with DREME E-value on the righ

CRX-dependent activated genes affected in opposite directions in developing CrxE80A and CrxK88N mutant retinas

(A) Heat map showing sample-wise Pearson correlations of the expression of all CRX-dependent activated genes between P10 WT and HD mutant mouse retinas in this study (rows) with post-natal WT retinas from age P3 to P21 (columns, data from GSE87064). (B) Heat map showing the expression changes of DEGs in CRX-dependent activated gene set in HD mutant mouse retinas at P10. (C-D) Heat maps showing expression changes of selected photoreceptor genes from Group 1 and Group 2. Color scale identical to (B).

E80A and K88N mutations affected the expression of CRX-dependent activated genes in opposite directions in a critical time window of photoreceptor differentiation

To understand how different CRX mutations affected gene expression at canonical and de novo binding sites and how these changes impair photoreceptor differentiation, we turned to bulk RNA-seq analysis from the developing retinas at P10. At P10, photoreceptors have started to differentiate, and the expression of many photoreceptor genes undergo exponential increase39, 40. To focus on the most relevant expression changes, we first defined a set of genes that most likely depend on CRX activity nearby for expression (Figure S4A, Methods). Briefly, we associated each CRX ChIP-seq peak to the nearest gene, filtered only genes with nearby CRX ChIP-seq peaks, and further narrowed the list of genes to those significantly down-regulated in the loss of function mutant CrxR90W/W. Gene ontology (GO) analysis confirmed that this putative CRX-dependent gene set is associated with biological processes related to photoreceptor development and functions (Figure S4B). This set of putative CRX-dependent genes also showed developmental dependent gain in expression, consistent with CRX’s primary function as a transcriptional activator (Figure S4C). As a control, CRX-independent genes were constitutively expressed and largely involved in general cellular processes (Figures S4D-S4F). Therefore, the CRX-dependent gene set comprises genes important for photoreceptor differentiation and functional maturation and are dependent on CRX for activation. We denote these genes as “CRX-dependent activated genes” (CRX-DAGs).

Next, we sought to understand how each mutation affected photoreceptor differentiation. One way of measuring the progression of photoreceptor differentiation is to determine the similarity in CRX-DAG expression in a given sample with that of known developmental ages in WT control animals. We thus performed sample-wise correlation of CRX-DAG expression obtained in our P10 samples with a previously published RNA-seq dataset of normal mouse retinal development (Figure 3A)39. As expected, our P10 WT sample showed strong correlations with all developmental ages in the published WT control dataset. A stronger correlation with early ages (P3, P7, P10) and a weaker correlation with later ages (P14, P21) is also an indication of ongoing photoreceptor differentiation at P10. Unlike the WT sample, CrxE80A/+ and CrxE80A/Asamples both showed a stronger correlation with later developmental ages (P14, P21) but a weaker correlation with earlier postnatal ages (P3, P7). Since the CRX-DAGs are normally developmentally upregulated, this shift in correlation towards later developmental ages suggested that these genes were prematurely upregulated in the P10 CrxE80A mutant retinas. In contrast, CrxK88N/+ and CrxK88N/N samples both showed a weaker correlation with all developmental ages when compared with WT samples in our dataset. This suggests that early photoreceptor differentiation was compromised in both CrxK88N mutants, consistent with their association with early-onset LCA23. Importantly, CrxR90W/W, also associated with LCA-like phenotype19, 22, displayed strong correlation with earlier ages (P3, P7) similar to WT, but only showed moderate correlation with later ages (P14, P21). This suggests loss of CRX function at canonical binding sites does not affect the initiation of photoreceptor differentiation, but WT CRX activity at these sites is required to sustain differentiation. Since CRX-DAG expression was more severely affected in CrxK88N mutants than in CrxR90W/W, the photoreceptor differentiation deficits seen in the CrxK88N mutants cannot be explained solely by the loss of regulatory activity at canonical CRX binding sites. Overall, our sample-wise correlation analysis with normal retinal development dataset suggests that E80A and K88N mutations affected the expression CRX-dependent activated genes in opposite directions, implicating novel and distinct pathogenic mechanisms from the loss-of-function R90W mutation.

CrxE80A retinas show up-regulation of rod genes but down-regulation of cone genes, underlying CoRD-like phenotype

Upon closer examination, we noted that not all CRX-DAGs were up-regulated in CrxE80A mutants (Figure 3B). Hierarchical clustering of all CRX-DAGs using expression changes revealed two major groups (Methods). In aggregate, when compared to WT, Group 1 genes were up-regulated in CrxE80A mutants while Group 2 genes were down-regulated. We noted genes indicative of the two photoreceptor subtypes, rods and cones, could partially define the two groups (Figures 3C and 3D). For example, Esrrb41 and Nrl42 in Group 1 are important regulators of rod differentiation. Other genes in Group 1 are components of the phototransduction cascade in rods, including Rcvrn43, Rho44, Gnat145, 46, Pde6g47, Abca448, 49, Gnb150, Rdh1251, Cngb152, Rp153, 54. Mis-regulations of many of these genes have been associated with diseases that affect rod development, function, and long-term survival. The increased activation of these genes likely underlies the stronger correlation with later developmental ages in CrxE80A+/and CrxE80A/A retinas (Figure 3A). In contrast, Group 2 genes, many down-regulated in CrxE80A mutants, were implicated in cone development and functions. For example, Gnat255, 56, Pde6c57, and Pde6h58, 59 all act in the cone phototransduction cascade. Mis-regulation of these genes has also been implicated in different retinal dystrophies that primarily affect cone photoreceptors. Comparison of ChIP-seq signal revealed that peaks associated with Group 2 genes showed lower occupancy compared to Group 1 genes in the CrxE80A/A retinas (Mann-Whitney U test p-value: 9.51e-07) while no difference was observed in WT retinas (Mann-Whitney U test p-value: 0.541), suggesting loss of CRX activity likely underlies the down-regulation of Group 2 genes in the CrxE80A mutants (Figure S5A). Collectively, the selective down-regulation of cone genes in Group 2 may explain the CoRD-like phenotype in adult CrxE80Amutant mice described later.

Additionally, we noticed that a subset of genes not affected in CrxR90W/W(CRX-independent genes) were also down-regulated in CrxE80A mutants (Figures S5B-S5D). Among these genes were transcription regulators important for early photoreceptor development, such as Ascl160, Rax61, Sall362, and Pias363. The down-regulation of these factors coincided with the up-regulation of mature rod genes in P10 CrxE80A retinas, suggesting that the E80A mutation might hamper the proper timing of photoreceptor differentiation.

CrxK88N retinas display greater reduction of rod and cone genes than the loss-of-function mutants

CrxK88N retinas had the most severe gene expression changes among all mutants with down-regulation of both Group 1 (rod) and Group 2 (cone) genes (Figures 3B-3D). The heterozygous CrxK88N/+ retina displayed a similar degree of expression reduction as homozygous CrxR90W/W, consistent with its association with dLCA23. Given the normal phenotype of heterozygous loss-of-function mutants - Crx+/-and CrxR90W/+19, these results suggest that mutant CRX K88N not only failed to activate WT target genes, but also functionally antagonized WT CRX regulatory activity in differentiating photoreceptors. This antagonism might be associated with ectopic CRX K88N activity when bound to regulatory regions with N88 HD DNA motifs (Figures 2A and 2E). In the absence of WT CRX, CrxK88N/N retina displayed a more severe expression reduction of CRX-DAG than CrxR90W/W, raising the possibility that CRX K88N also antagonized the activity of other transcriptional regulators important for photoreceptor differentiation. Supporting this possibility, a set of CRX-independent genes were also mis-regulated in both heterozygous and homozygous CrxK88N mutants. We noted that a number of these down-regulated genes are also involved in photoreceptor functional development (Figures S5B, S5E, and S5F). Overall, CRX K88N is associated with greater gene expression changes than other CRX HD mutants, which is likely attributed to ectopic regulatory activity.

Both E80A and K88N mutants show compromised rod/cone terminal differentiation in young adults

Since Crx+/- and CrxR90W/+ mutant mouse models showed a late-time recovery in photoreceptor gene expression and function20, we sought to determine the degree of photoreceptor differentiation in CrxE80A and CrxK88N mutants at P21. At this age, the normal retina has largely completed terminal differentiation with photoreceptor gene expression reaching a plateau64. However, when P21 CRX HD mutant retinas were examined for the expression of genes under the GO term detection of light stimulus (GO:0009583), which comprises genes in both rod and cone phototransduction cascades, many genes failed to reach WT levels, despite variable degrees of impact across different HD mutants (Figure 4A). CrxE80A mutants, in contrast to the increased rod gene expression at P10, displayed a deficiency in both cone and rod phototransduction genes at P21 (Figure 4B, Supplementary Table 6). This suggests that mutant CRX E80A transcriptional activity fails to sustain photoreceptor terminal differentiation and ultimately results in non-functional and severely affected photoreceptors. In comparison, CrxK88N mutants showed severely reduced expression of rod/cone phototransduction genes at both P10 and P21 (Figures 3B-3D and 4B).

Photoreceptor genes important for phototransduction are down-regulated in all HD mutants

(A) Box plot showing that genes in the detection of light stimulus GO term were down-regulated and affected to various degrees in different adult (P21) HD mutant mouse retinas. (B) Heat map showing that expression of both cone and rod phototransduction genes were down-regulated in adult (P21) HD mutant mouse retinas. Annotation of rod and cone enrichment of each gene is in Supplementary Table S6. See Supplementary Figure S6 for the developmental expression dynamics of these genes.

To assess retinal morphology and photoreceptor subtype-specific gene expression at the cellular level, we performed immunohistochemistry analysis on P21 retinal sections. In WT animals, a hallmark of photoreceptor maturation is the outgrowth of photoreceptor outer segments (OS) filled with proteins necessary for the phototransduction. We thus performed hematoxylin and eosin (H&E) staining on P21 sagittal retinal sections to visualize changes in retinal layer organization, focusing on photoreceptor layers – outer nuclear layer (ONL) and OS. Compared to the well-organized ONL in WT retinas (Figure 5A), all mutants showed variable degrees of ONL disorganization, forming waves, whorls, and rosettes (Figures 5B-5E). The ONL disorganization was more severe in homozygotes than in heterozygotes for both mutations and CrxK88Nmutants were more severely affected than CrxE80A mutants. Photoreceptor OS layer was formed in the CrxE80A/+ retinas, but absent in CrxE80A/A, CrxK88N/+, and CrxK88N/N mutant retinas. Inner retinal layers, including inner plexiform layer (IPL) and ganglion cell layer (GCL) were not as severely affected as the outer retinal layers, supporting a model that the mutant morphological abnormalities largely originated from the diseased photoreceptors. These morphological abnormalities were distinct from the degenerative phenotypes of other Crx mutant models reported previously19, 65, 66.

Only CrxE80A/+ retinas maintain photoreceptor OS and residual rod ERG response

(A-E) Hematoxylin-eosin (H&E) staining of P21 retina sections show that photoreceptor OS layer is absent in all mutant retinas except CrxE80A/+. OS: outer segment; ONL: outer nuclear layer; INL: inner nuclear layer; IPL: inner plexiform layer; GCL: ganglion cell layer. Scale bar, 100µm. (F-J) Rhodopsin (RHO, red) immunostaining is present in CrxE80A/+, CrxE80A/A, and CrxK88N/+ retinas and absent in CrxK88N/N retina. Cone arrestin (mCAR, green) immunostaining is absent in all mutant retinas. Nuclei were visualized by DAPI staining (Blue). Scale bar, 100µm. (K-M) The electroretinogram responses (ERG) recorded from 1-month mice. Error bars represent the standard error of the mean (SEM, n ≥ 4). p-value: Two-way ANOVA and Tukey’s multiple comparisons. ****: p ≤ 0.0001. ns: >0.05.

Next, we sought to determine the expression of the rod-specific visual pigment rhodopsin (RHO) and cone arrestin (mCAR) in the P21 mouse retinas. In WT retinas, RHO is trafficked to the rod OS while mCAR is present in the cone OS and IS (inner segment), cell body, and synaptic terminals (Figure 5F). Unlike WT retina, all mutants lacked mCAR immunoreactivity (Figures 5G-5J), consistent with the loss of cone gene expression shown by RNAseq (Figure 4B). In CrxE80A/+ retinas, RHO staining was localized to the OS layer; in CrxE80A/A and CrxK88N/+retinas, positive RHO staining was observed within the whorls and rosettes; in CrxK88N/N mutant retinas, RHO staining was completely absent. Importantly, we did not observe mis-localized RHO staining in the inner retinal layers (INL) suggesting that the developmental programs of other retinal cell types were not directly affected by E80A or K88N mutation. Overall, abnormalities in the cone/rod gene expression matched the corresponding human disease diagnosis21, 23, and the phenotypic severity correlated with the degree of mis-regulation of CRX target genes in the corresponding RNAseq dataset. Thus, these results support a model that CRX HD mutation-mediated mis-regulation of gene expression disrupts photoreceptor terminal differentiation and leads to defects in retinal layer organization and OS formation.

E80A and K88N mouse models show visual function deficits that recapitulate human diseases

To understand the consequences of disrupted photoreceptor differentiation on visual function, we measured electroretinogram (ERG) responses to light stimuli for WT and mutant mice at one month of age (Figures 5K-M). In response to incremental changes of light intensities, WT animals showed corresponding amplitude increases in dark adapted A-waves (rod signals) and B-waves (rod-evoked bipolar cell signals), as well as in light-adapted B-waves (cone-evoked bipolar cell signals). The three severe mutants, CrxE80A/A, CrxK88N/+, and CrxK88N/N had no detectable dark-adapted or light-adapted ERG responses, suggesting that these mice have no rod or cone function and are blind at young ages. The null ERG phenotype of the CrxK88N animals is consistent with the clinical LCA phenotype in humans23. In contrast, CrxE80A/+ animals retained partial rod ERG responses as indicated by the reduced A-wave and B-wave amplitudes (Figures 5K and 5L). Yet, CrxE80A/+ animals had no detectable cone ERG responses, which is consistent with the CoRD clinical phenotype in humans21 (Figure 5M). Taken together, the visual function impairment in each CRX HD mutant model, coincided with the morphological and molecular changes, suggesting that CrxE80A and CrxK88Nmouse models recapitulate the corresponding human diseases.

CRX E80A has increased transactivation activity and leads to precocious differentiation in CrxE80A retinas

Lastly, we asked what might be the molecular mechanism that causes the mis-regulation of photoreceptor genes in the mutant retinas. Previous studies have established that reporter assays with the rhodopsin promoter in HEK293T cells can measure changes of CRX transactivation activity and inform the mechanisms by which photoreceptor genes are mis-regulated in CRX mutant retinas18. We thus tested the transactivation activity of the three CRX HD mutants on the pRho-Luc reporter in HEK293T cells (Figure 6A, Methods). Consistent with published studies, R90W mutant had significantly reduced activity compared to WT CRX18. K88N mutant showed a similarly reduced activity as R90W, consistent with K88N’s loss of binding at canonical CRX sites and failure to activate CRX-DAGs in vivo (Figures 2A, 2B, 3A-3C). In contrast, E80A mutant, which binds to canonical CRX sties, showed significantly increased transactivation activity on the Rho promoter. This hyperactivity of E80A protein at Rho promoter correlates with the upregulation of CRX-DAGs in the mutant retinas at P10 (Figures 3A-3C).

CRX E80A hyperactivity underlies precocious photoreceptor differentiation in CrxE80A retinas.

(A) Boxplot showing luciferase reporter activities of different CRX variants. P-values for one-way ANOVA with Turkey honestly significant difference (HSD) test are indicated. p-value: ****: ≤0.0001, ***: ≤0.001, ns: >0.05. (B-D) Rhodopsin (RHO, green) immunostaining is absent in P3 WT retina but detected in CrxE80A/+ and CrxE80A/A retinas. Nuclei are visualized by DAPI staining (Blue). Arrow indicates the sporadic RHO staining in CrxE80A/+ sample. ONBL: outer neuroblast layer; GCL: ganglion cell layer. Scale bar, 100µm.

A transition to the next developmental stage usually requires the expression of important developmental genes passing an abundance threshold. Based on the hyperactivity model, photoreceptor genes are activated stronger in CrxE80A retinas and thus could reach the abundance threshold earlier. To determine the consequences of E80A hyperactivity on photoreceptor differentiation timing, we compared RHO protein expression during early postnatal retinal development using retinal section immunostaining (Figures 6B-6D). In WT retinas, most rods were born by P3 but had not differentiated (Figure 6B). Previous studies showed that RHO proteins were detected by IHC starting around P7 in WT retinas64. In comparison, both CrxE80A/+ and CrxE80A/A retinas showed positive RHO staining at P3 (Figures 6C and 6D). RHO+ cells were largely seen in the outer portion of the ONBL layers in CrxE80A/+ retinas, and strikingly spread throughout the large presumptive ONL layers in CrxE80A/A retinas. The detection of RHO protein in P3 CrxE80A mutant retinas indicates that photoreceptor differentiation program was precociously activated. Taken together, our results support a model that E80A and K88N mutations each perturbs CRX regulatory activity in a unique way, causes photoreceptor differentiation defects, and ultimately leads to distinct dominant disease phenotype that recapitulates human diseases.

Discussion

Through molecular characterization of mutant proteins, transcriptome, and cellular profiling of developing mutant mouse retinas and ERG testing of adult retinas, we have identified two novel pathogenic mechanisms of CRX HD mutations, E80A and K88N, that are associated with dominant CoRD and dominant LCA in human21, 22. Distinct from the previously characterized loss-of-function R90W mutation19, 20, E80A and K88N mutations produce altered CRX proteins with gain of regulatory functions - CRX E80A is associated with increased transcriptional activity and CRX K88N has altered DNA binding specificity. Both CRX E80A and CRX K88N proteins impair photoreceptor gene expression, development and produce structural and functional deficits in knock-in mouse models, recapitulating human diseases65 (Figure 7). Thus, both target specificity and regulatory activity precision at the canonical CRX targets are essential for proper photoreceptor development and functional maturation.

Missense mutations in CRX HD affect photoreceptor gene expression and leads to distinct retinal disease phenotypes through gain- and loss-of-function mechanisms.

Although associated with distinct disease phenotypes, the E80A, K88N, and R90W mutations are located very close to each other in the CRX HD recognition helix. Extensive biochemical and structural studies on the HD-DNA complexes afford important insights into how these mutations could affect CRX HD-DNA interactions differently. Most distinctively, CRX K88 residue, at HD50 position, is the major contributor to HD-DNA binding specificity, with lysine making favorable interactions with both guanines in the CRX consensus TAATCC binding site67, 68. It is thus expected that K88N mutation drastically changes CRX DNA binding preference at the 3’ end of the HD motif, reminiscent of previous findings on novel HD DNA-binding specificity using bacterial one-hybrid (B1H) systems31, 69. Supporting the importance of K88 residue mediated CRX target specificity in regulating photoreceptor development, CrxK88Nretinas show more severe perturbations in photoreceptor gene expression and development than in loss-of-function mutant CrxR90W/W. CrxK88N/+and CrxK88N/N mice display the most severe photoreceptor morphological deficits observed in any Crx mouse models and show absence of visual functions in young adults. Thus, CRX target specificity is critical for photoreceptor development fidelity.

In developing WT mouse retinas, the HD motif preferred by HD TFs with a glutamine (Q) at HD50 position encodes quantitatively different activity than the CRX consensus suggesting functional difference between HD binding site variants61, 70. It is likely that the severe CrxK88N phenotypes are attributed to both diminished activity at canonical CRX motifs and ectopic binding and transcriptional activity at N88 HD motifs. Since a functional copy of WT CRX is retained in CrxK88N/+ retinas, the lack of WT activity alone cannot explain the severe developmental deficits. Alternatively, these results suggest involvement of additional regulatory mechanisms: CRX K88N activity at N88 HD motifs might (1) ectopically activate genes whose expression prevents the progression of development or inactivate genes required for development; (2) interfere with other HD TFs that also recognize N88 HD motifs; (3) lead to epigenetic alterations that antagonize normal CRX functions. Many other HD containing TFs are expressed in developing mouse retina, including OTX2, RAX, VSX2, PAX6, SIX3/6, and LHX family6. Different from CRX, these HD TFs are essential for gene regulation in retinal progenitor cells and/or in other retinal cell lineages. Alteration of CRX DNA binding specificity could mis-regulate genes originally targeted by these HD TFs and lead to severe perturbations in the retinal gene regulatory networks (GRNs). To date, most studies have focused on CRX activity at cis-regulatory sequences enriched for the WT CRX consensus motifs71, 72. Systematic comparison of regulatory activity at N88 HD motifs and WT consensus in the context of photoreceptor development in both WT and mutant retinas would be needed to substantiate the impact of mutant CRX K88N activity at different HD motifs. These experiments will also help clarify the pathogenic mechanisms in the CrxK88Nmodels and extend our knowledge of CRX HD mediated regulatory grammar during photoreceptor development.

Different from CRX K88, the E80 and R90 residues, although the most common residues at HD42 and HD52 positions respectively, do not contact DNA directly and thus lacked in-depth investigations in prior studies. CRX R90 residue has been suggested to confer additional stability for the homeodomain fold besides the core residues and make contacts with the DNA backbone through bases in the TAAT core motif67, 68. Substitution of the basic R90 residue with a bulky, neutral tryptophan (W) potentially reduces overall CRX HD stability which in term reduces CRX HD DNA-binding affinity without affecting its binding preference. The potential reduction in CRX HD-DNA complex stability is in line with our observation that R90W mutation abolishes CRX binding across the genome resulting in global loss of CRX target gene activation. It also explains the association of CRX R90W mutation with recessive loss-of-function LCA phenotypes in human and mouse22, 65.

Structural studies suggest that CRX E80 residue plays a role in stabilizing the HD-DNA binding complex through intramolecular interactions with other HD residues29, 68, yet functional validations await further experiments. E80A mutation, replacing glutamic acid (E), which is acidic and polar, with alanine (A), which is neutral and non-polar, could render HD-DNA interactions more promiscuous as reflected in overall reduced magnitude of CRX E80A HD specificity (Figure 1I). Regulatory sequences of many photoreceptor genes contain both consensus and non-consensus CRX motifs13, 14, 18, 35. It is likely that the more promiscuous CRX E80A-DNA interaction increases the likelihood of non-consensus CRX HD motifs being bound and activated resulting in overall increased transcriptional output (hyperactivity) as seen both in luciferase assays and in developing CrxE80A mouse retinas. The promiscuous TF-DNA-binding associated hyperactivity phenomenon has also been observed in a dominant disease mouse model harboring a missense mutation in the zinc finger TF Krüppel-like factor-1 (KLF1)73. Yet, in adult CrxE80A retinas, photoreceptor terminal differentiation is impaired, resulting in disrupted retinal morphology and defective visual functions. Photoreceptor differentiation is programmed via sequential and concerted gene expression programs within a defined time window74, 75. One explanation for adult CrxE80A phenotypes is that CRX E80A hyperactivity precociously activates later stage genes in the absence of proper nuclear context and/or subcellular structures, which in turn negatively impacts early events in photoreceptor differentiation. These observations underscore the importance of precisely tuned CRX-mediated transcriptional activity during photoreceptor development.

Although a global increase in expression is expected in CrxE80A retinas based on the hyperactivity model, a subset of CRX dependent activated genes implicated in cone photoreceptor development and functions is down-regulated in both differentiating and mature mutant retinas. While cones undergo terminal differentiation to develop into cone subtypes – M- or S-cones – in a similar postnatal window as rods, they were born prenatally in mice within an earlier time window than rods. At an early postnatal age, cells expressing RXRγ, a ligand-dependent nuclear hormone receptor normally expressed in developing cones60, 76–78, were observed in CrxE80A retinas, suggesting cone photoreceptors were born in these mutant retinas (Figure S7A). One model for lack of cone markers in adult CrxE80A retinas is that CRX E80A improperly activates later stage cone genes at a much earlier time window, disrupting cone terminal differentiation. Supporting this model, CRX E80A also hyperactivates the S-cone opsin promoter (Figure S7B). An alternative model is that cones might be more sensitive to perturbations in CRX activity. It is known that cones depend on a different repertoire of TFs than rods for subtype terminal differentiation60. It is possible that cone TFs respond differently to mutant CRX E80A hyperactivity, leading to the distinct expression changes in CrxE80A mutant retains. It is important to note that different point mutations at CRX E80 residue have been reported in dominant CoRD cases (ClinVar VCV000865803.1, VCV000007416.7, VCV000099599.6), emphasizing the importance of residue CRX E80 in regulating cone photoreceptor development. Since cones only make up a very small portion (3%) of photoreceptors in mouse retinas79, quantitative characterization of CRX E80A activity in a cone dominant retina will be necessary to understand its selective effect on the cone differentiation program and help elucidate WT CRX regulatory principles in early photoreceptor development.

Given that the spatial structures and HD-DNA contact models of HD proteins are evolutionarily conserved, our study of CRX provides valuable molecular insights for HD mutations implicated in other diseases. For example, p.E79K substitution (corresponds to CRX E80) in OTX2 HD is associated with dominant early-onset retinal dystrophy80, heterozygous p.R89G (corresponds to CRX R90) mutation in OTX2 HD causes severe ocular malformations81, and missense mutations of the CRX K88 and R90 homologous residues in PITX2 HD are associated with dominant Rieger syndrome82. It is likely that these mutations affect HD activity in similar ways as observed in CRX, and the exact disease manifestation is determined by cell-type or tissue specific mechanisms. The retina is readily accessible, and a broad range of molecular tools are available for ex vivo and in vivo manipulations. We believe that CRX is an ideal model to study the pathogenic mechanisms of HD mutations and to test therapeutic regimens, which would ultimately benefit the study of HD TFs and their associated diseases in other tissues and organs.

One limitation of this work is that effects of E80A and K88N mutations on CRX HD-DNA interactions have been evaluated at monomeric HD motifs and with homogenous protein species both in vitro and in vivo. Further evaluation of WT and mutant CRX binding at dimeric motifs will be desirable, since selected dimeric HD motifs are known to mediate HD TF interactions to ensure gene expression fidelity during development83, 84. Relatedly, we also need to address how CRX WT and mutant E80A or K88N proteins interact at HD binding motifs – whether they cooperate or compete with each other, whether these interactions are HD motif sequence-dependent, and how gene expression is impacted by CRX cooperativity or competition. While CRX HD mediates both TF-DNA interactions and protein-protein interactions, evaluation of how E80A and K88N mutations impact CRX interaction with other important photoreceptor TFs and how perturbations in these interactions lead to disease phenotypes warrant further study.

Collectively, our findings support a unifying model in which precise CRX interaction with cis-regulatory sequences is essential for gene expression and functional maturation during photoreceptor development. Disease-associated mutations in CRX have been classified into two main groups – insertion/deletion-derived frameshift mutations in the activation domain (AD) and missense mutations in the homeodomain65. Prior biochemical and mouse model studies of the first group have established that AD-truncated mutant proteins abolish CRX transcriptional activity and functionally interfere with the WT allele. As a result, the mutant retinas fail to activate or maintain robust cone/rod gene expression, resulting in incomplete photoreceptor differentiation and ultimately rapid degeneration of immature photoreceptors16, 19. In this study, we demonstrate that missense mutations in the CRX HD, by either a loss- or gain-of-function mechanism, alter CRX target specificity and/or CRX transactivation activity. These biochemical property changes impair CRX-mediated transcriptional regulation in vivo and lead to distinct morphological and functional deficits (Figures 7A-7D). Despite the difference in molecular mechanisms, both CrxE80A and CrxK88N mouse models develop whorls and rosettes in the ONL by P21, which are not observed in degenerative CRX mouse models19, suggesting distinct pathogenic mechanisms. Future cellular biology studies are needed to understand the formation mechanisms of these unique cellular phenotypes (ONL disorganization) and their impacts on the function and survival of photoreceptors and other retinal cell types over development.

Our study here also emphasizes the importance of tailoring gene therapy regimens to tackle individual pathogenic mechanisms. For instance, while supplementing WT CRX might be sufficient to rescue a hypomorphic/loss-of-function mutant, simultaneous elimination of a gain-of-function CRX product would be necessary to rescue dominant mutants, as exemplified in a recent report of allele-specific gene editing to rescue dominant CRX-associated LCA7 phenotypes in a retinal organoid model85. We believe that this principle also applies to other dominant neurological diseases. Additionally, with the refinement of the CRX mechanistic model, when new disease mutations are identified, genetic counsellors can now provide more informed predictions of disease progression and future visual deficits. This information is important for individuals to be psychologically prepared and seek necessary assistance to improve their quality of life.

Author contributions

S. Chen conceived and supervised the study. S. Chen, P. Ruzycki, C. Sun and Y. Zheng designed the experiments. Y. Zheng performed Spec-seq and luciferase experiments; X. Zhang performed CRX ChIP-seq experiments; C. Sun performed immunochemistry and ERG experiments. Y. Zheng analyzed Spec-seq, CRX ChIP-seq, RNA-seq, and luciferase data; P. Ruzycki assisted in CRX ChIP-seq and RNA-seq data analysis; C. Sun analyzed immunochemistry and ERG data. Y. Zheng wrote the original draft; S. Chen, P. Ruzycki, C. Sun, and Y. Zheng revised the manuscript. All authors read and approved the final manuscript.

Supporting information

eLife Assessment

This manuscript is of interest to readers in the field of neural development and neurodegeneration. The study is important as it examines two disease-causing mutations within the homeodomain transcription factor Cone-Rod Homeobox (CRX) that causes retinopathy in humans. The data are solid and the work contributes to our understanding of the underlying pathogenetic mechanisms.

Reviewer #1 (Public Review):

The manuscript by Zheng et al. examined the disease-causing mechanisms of two missense mutations within the homeodomain (HD) of CRX protein. Both mutations were found in humans and can produce severe dominant retinopathy. The authors investigated the two CRX HD mutants via in vitro DNA-binding assay (Spec-seq), in vivo chromatin-binding assay (ChIP-seq), in vivo expression assay of downstream target genes (RNA-seq), and retinal histological and functional assays. They concluded that p.E80A increased the transactivation activity of CRX and resulted in precocious photoreceptor differentiation, whereas p.K88N significantly changed the binding specificity of CRX and led to defects in photoreceptor differentiation and maintenance. The authors performed a significant amount of analyses. The claims are sufficiently supported by the data. The results not only uncovered the underlying disease-causing mechanisms, but also can significantly improve our understanding of the interaction between HD-TF and DNA during development.

Reviewer #2 (Public Review):

Zheng et al., investigated the molecular and functional mechanisms of two homeodomain missense mutations causing human retinal photoreceptor degeneration diseases in photoreceptor development regulated by the CRX transcription factor. They analyzed the E80A mutation associated with dominant cone-rod dystrophy (CRD) and the K88N mutation associated with dominant Leber Congenital Amaurosis (LCA). The authors found that E80A CRX binds to the same target DNA sites as WT CRX, but the binding specificity of K88N CRX is altered from that of WT in an in vitro assay. They generated Crx(E80A) and Crx(K88N) KI mice and performed ChIP assay and observed that K88N CRX binds to novel genomic regions from the WT-binding sites, while E80A binds to the WT sites. In addition, using the KI mice, they found that E80A and K88N differently affect the expression of Crx target genes. This study is well executed with proper and solid methodologies, and the manuscript is clearly written. This study gives us the insights into how single missense CRX mutations lead to different types of human retinal photoreceptor degeneration diseases.

Overall, the authors have significantly improved the manuscript, but there is still an unclarified point. In response to the inquiry in the initial review on how extent E80A KI mice function as a pathological model of dominant CoRD, the authors add data (Figures S7) and described the sixth section in the discussion. However, the authors mentioned that it is technically too challenging because of a small number of cones. The point is not clear to me, but it is possible to analyze cone differentiation and degeneration by immunostaining at multiple stages even though cone number is small. Cone arrestin and S- and M-opsins become positive at early postnatal stages in the mouse retina. Cone arrestin seems earlier than cone opsins. Cones seem born by detecting RXRg at P0, but are cone arrestin and/or cone opsins expressed in early postnatal E80A/+ retina? If positive, how about an apoptosis marker? If negative, it seems to be a cone development phenotype rather than cone degeneration phenotype. If so, authors should modify the expression to say that the E80A retina underlies CoRD-like phenotype. It seems an overstatement.

Author Response:

The following is the authors’ response to the previous reviews

Thank you for sending our revised manuscript for review and the positive editorial comments. On behalf of all authors, I would like to, again, thank the reviewers for their critical reading of our revised manuscript and for providing further suggestions. We have revised the introduction and discussion sections to specifically address the comments made by Reviewer #2. Please see below for detailed explanations.

Reviewer #2 (Public Review):

Overall, the authors have significantly improved the manuscript, but there is still an unclarified point. In response to the inquiry in the initial review on how extent E80A KI mice function as a pathological model of dominant CoRD, the authors add data (Figures S7) and described the sixth section in the discussion. However, the authors mentioned that it is technically too challenging because of a small number of cones. The point is not clear to me, but it is possible to analyze cone differentiation and degeneration by immunostaining at multiple stages even though cone number is small. Cone arrestin and S- and M-opsins become positive at early postnatal stages in the mouse retina. Cone arrestin seems earlier than cone opsins. Cones seem born by detecting RXRg at P0, but are cone arrestin and/or cone opsins expressed in early postnatal E80A/+ retina? If positive, how about an apoptosis marker? If negative, it seems to be a cone development phenotype rather than cone degeneration phenotype. If so, authors should modify the expression to say that the E80A retina underlies CoRD-like phenotype. It seems an overstatement.

We greatly appreciate Reviewer 2’s suggestions on further investigating cone photoreceptor phenotypes in the CRX E80A KI mouse model. All the points raised deserve a comprehensive and in-depth study. However, the focus of the current manuscript is to establish a general framework for understanding different missense mutations in homeodomain TFs beyond CRX. We believe that a separate and dedicated study is more appropriate to detail the quantitative molecular and cellular mechanisms of CRX E80A dysfunction in cone and rod photoreceptors, as stated in the last sentence of discussion section paragraph 6: “… quantitative characterization of CRX E80A molecular functions in a cone dominant retina warrants further study to understand its selective effect on the cone differentiation program and help elucidate WT CRX regulatory principles in early photoreceptor development.”.

Clinical diagnosis of cone-rod dystrophy (CoRD) is largely based on functional deficits of cones and rods. 1-month electroretinogram (ERG) (Figures 5K-M) shows no cone-mediated light responses and reduced rod functions in CrxE80A/+ mouse. These ERG deficits in the CRX E80A KI mouse model are in agreement with CoRD characteristics. Thus, it is reasonable to say that CRX E80A KI retina phenotype resembles CoRD phenotype.

Reviewer #2 (Recommendations For The Authors):

As a minor comment, in page 8, second section, "Previous studies have demonstrated the CRX is activated shortly after cell cycle exit in retinal progenitor cells fated to be photoreceptor.", the authors cited refs 66 and 67, which were in 2105 and 2016. However, it was demonstrated in the paper of J. Neurosci.31(46), 16792-807, 2011, Figure 1. The authors need to be scientifically fair to cite the JN 2011 paper.

In response to this comment above, the authors cited the JN 2011 paper in a modified sentence of "Animal studies have demonstrated that Crx is first expressed in post-mitotic photoreceptor precursors and maintained throughout life (Refs.13-15)", moved from the discussion to the introduction. To my knowledge, the JN2011 (new Ref 15) is the first study directly demonstrated that Crx begins to be expressed shortly after cell cycle exit of retinal progenitor cells. Refs. 13 and 14 showed Crx expression in adult stage photoreceptors but did not directly demonstrate the Crx expression in post-mitotic photoreceptor precursors. To be scientifically precise, the references should be cited as "Animal studies have demonstrated that Crx is first expressed in post-mitotic photoreceptor precursors (Ref. 15) and maintained throughout life (Refs.13 and 14)".” Thanks to the reviewer for the precise instruction. We have adjusted the reference order as follows: “Animal studies have demonstrated that Crx is first expressed in post-mitotic photoreceptor precursors13 and maintained throughout life14,15.”, where JN2011 paper is reference 13.

Version 3

Abstract

Summary

Homeodomain transcription factors (HD TFs) are instrumental to vertebrate development. Mutations in HD TFs have been linked to human diseases, but their pathogenic mechanisms remain elusive. Here we use Cone-Rod Homeobox (CRX) as a model to decipher the disease-causing mechanisms of two HD mutations, p.E80A and p.K88N, that produce severe dominant retinopathies. Through integrated analysis of molecular and functional evidence in vitro and in knock-in mouse models, we uncover two novel gain-of-function mechanisms: p.E80A increases CRX-mediated transactivation of canonical CRX target genes in developing photoreceptors; p.K88N alters CRX DNA-binding specificity resulting in binding at ectopic sites and severe perturbation of CRX target gene expression. Both mechanisms produce novel retinal morphological defects and hinder photoreceptor maturation distinct from loss-of-function models. This study reveals the distinct roles of E80 and K88 residues in CRX HD regulatory functions and emphasizes the importance of transcriptional precision in normal development.

Introduction

Homeodomain transcription factors (HD TFs) play a fundamental role in vertebrate development. Members of the HD TF family are characterized by the presence of a highly conserved 60 amino acid helix-turn-helix DNA-binding domain known as the homeodomain (HD). The HD is one of the most studied eukaryotic DNA-binding motifs since its discovery in Drosophila homeotic transformations1,2. Hundreds of HD TFs have subsequently been documented in regulating gene expression programs important for body plan specification, pattern formation and cell fate determination1,2. Mutations in HD TFs have been linked to many human diseases, including neuropsychiatric and neurodegenerative conditions3,4. Although significant progress has been made in understanding HD-DNA interactions, uncovering the pathogenetic mechanisms of disease-causing missense mutations in HD have proven challenging.

The retina has long been used as a model system to study the role of HD TFs in normal central nervous system (CNS) development and in neurological diseases5. During retinogenesis, HD TFs play essential roles in the patterning of neuroepithelium, specification of retinal progenitors and differentiation of all retinal cell classes that derive from a common progenitor6. Importantly, many HD TFs are shared between the brain and the retina during development and mutations in these TFs can lead to disease manifestation in both tissues7–12. The accessibility and wealth of available molecular tools makes the retina a valuable tool to decipher the pathogenic mechanisms of HD TF mutations associated with neurological diseases.

Here, we study CRX, a HD TF essential for photoreceptor cells in the retina, as a model to understand how single amino acid substitutions in the HD impact TF functions and cause blinding diseases. Photoreceptors are the most numerous neurons in the retina and are specialized to sense light and initiate vision through a process called phototransduction. Animal studies have demonstrated that Crx is first expressed in post-mitotic photoreceptor precursors13 and maintained throughout life14,15. Loss of CRX results in impaired photoreceptor gene expression, failure of maturation and rapid degeneration of immature, non-functional photoreceptors16. Protein-coding sequence variants in human CRX have been associated with inherited retinal diseases (IRDs) that affect photoreceptors: Leber congenital amaurosis (LCA), cone-rod dystrophy (CoRD), and retinitis pigmentosa (RP) (OMIM:602225). However, the recessive phenotype observed in Crx knockout mouse models fails to recapitulate many dominant human CRX mutations that arise de novo16.

CRX contains two functional domains – the N-terminal HD and C-terminal activation domain (AD) (Figure 1A), both are required for proper activation of target genes and maintenance of normal Crx mRNA transcript abundance17,18. To understand how CRX HD mutations cause diseases, we have previously reported a mutation knock-in mouse model carrying a hypomorphic mutation p.R90W (R90W) in CRX HD19,20. We found that R90W mutation produces a recessive loss-of-function phenotype very similar to that of Crx knockout mice.

Disease associated missense mutations altered CRX HD DNA binding specificity.

(A) Diagram of CRX functional domains: Homeodomain (HD) for DNA-binding and Activation Domain (AD) for target gene transactivation. The three missense mutations in this study are located at the C-terminus of CRX HD and associated with different retinal diseases in human. Number in the parenthesis denotes the CRX HD position of the corresponding mutated residue. (B) Alignments of HD recognition helix sequences for the indicated HD proteins for which HD missense mutations have been associated with inherited diseases. Accession numbers can be found in Supplementary Table S1. Missense variants in this study (highlighted) are located at highly conserved residues across species and different HD TFs. (C) Spec-seq experimental workflow (Methods). (D) Spec-seq library design of monomeric HD binding sites. (E) EMSA gel images of Spec-seq experiments with different CRX HD species. Bx: Bound. B-: Unbound. (F) Relative binding energy comparison from two different experiments with WT HD. (G) Binding energy model for WT CRX HD. (H-J) Relative binding energy comparison between WT HD and R90W HD (H), E80A HD (I), or K88N HD (J). Consensus sequence is defined to have relative binding energy of 0kT (TAATCC for WT, R90W and E80A, TAATTA for K88N). The identity line is represented in grey dash. The orange dashed line shows the best linear fit to the data. (K-M) Binding energy models for R90W HD (K), E80A HD (L), and K88N HD (M). Only sequence variants within two mismatches to the corresponding consensus sequences were used to generate binding models. Negative binding energy is plotted such that bases above the x-axis are preferred bases and bases below the x-axis are unfavorable bases. Constant bases (TAA) carried no information are drawn at arbitrary height in grey.

Intriguingly, several missense mutations within the same HD recognition helix as R90W, including p.E80A (E80A) and p.K88N (K88N), are linked to severe dominant IRDs21–23 (Figures 1A and 1B). Here, we utilized a multi-omics approach to investigate the functional consequences of the E80A and K88N mutations on CRX regulatory activities and photoreceptor development (Figure S1). Comparison of the in vitro HD-DNA binding models of CRX and disease variants generated by Spec-seq revealed unique specificity changes of each mutant protein. Introduction of each mutation into the endogenous Crx locus generated knock-in mouse models CrxE80A and CrxK88N that reproduced dCoRD- and dLCA-like phenotypes. ChIP-seq analysis of CRX binding in vivo revealed mutation-specific changes in CRX targetome, consistent with mutation-specific DNA binding changes in vitro. Retinal RNA-seq analysis uncovered two distinct mechanisms by which the two HD missense mutations contribute to altered gene expression programs during photoreceptor differentiation and maturation. Our results highlight the importance of residues E80 and K88 in CRX-mediated transcriptional regulation during photoreceptor development and the diverse mechanisms by which HD missense mutations can affect TF functions and lead to severe dominant neurological diseases.

Results

K88N but not E80A mutation alters CRX HD DNA-binding specificity in vitro

CRX belongs to the paired-like HD TF family that recognize a 6-bp DNA motif in a stereotypic way24–30. Extensive studies of the HD have revealed a canonical HD-DNA recognition model where recognition of the 3’ region (bases 4-6) of the HD DNA binding site is mediated by specificity determinants within the conserved HD recognition helix, corresponding to CRX residues 80-9624–27,30,31 (Figure 1B). In particular, HD residue 50, equivalent to CRX K88 residue (Figure 1A), is the major specificity determinant in paired-like HD TF-DNA interactions25,26. Since the three disease-associated HD missense mutations, E80A, K88N and R90W, are located within the CRX HD recognition helix, we wondered if these mutations change CRX HD DNA-binding specificity.

We adapted a high-throughput in vitro assay, Spec-seq, that determines protein-DNA-binding specificity by sequencing32–34. Spec-seq was developed based on the traditional electrophoretic mobility shift assay (EMSA) to measure protein-DNA interactions. Spec-seq allows us to measure the relative binding affinities (i.e., specificity) for a library of HD binding motifs in parallel and generate quantitative binding models for different CRX HDs (Figures 1C). Based on the HD-DNA interaction model, we designed and tested a Spec-seq library containing all possible monomeric HD motifs (TAANNN) (Figure 1D).

We first obtained the wild-type (WT) CRX HD DNA binding model with Spec-seq using bacterially-expressed and affinity-purified HD peptides (Figures 1E-1G, Methods). Relative binding energies of CRX WT HD from two experiments showed strong correlation (r: 0.984) and noise level (0.114 kT) within the expected range in typical Spec-seq data (Figure 1F). Binding energy model of WT HD was then generated by applying multiple linear regressions on the relative binding energies of all sequences within two base-pair mismatches to the WT CRX consensus (TAATCC)14 (Figure 1G, Methods). A clear preference for CC bases at the 3’ end of the motif is consistent with known CRX binding preference in vitro and in vivo35,36.

We next sought to understand how disease mutations affect CRX DNA-binding specificity. We purified all mutant HD peptides following the same protocol as WT HD peptides and verified their DNA binding (Figures S2A-S2D)18. Comparison of the relative binding energies between each pair of mutant and WT HD revealed distinct effects (Figures 1H-1J). By definition, the consensus DNA binding motif of a testing peptide has a relative binding energy of 0kT. The relative binding energy difference between nucleotide variants and the consensus motif correlates with the DNA binding specificity of the testing peptide. We found that R90W HD and E80A HD both prefer the same consensus motif as WT (Figures 1K and 1L). R90W HD bound with slightly higher specificity than WT, as demonstrated by most data points falling above the identity line (Figures 1H and 1K), suggesting that R90W HD is more sensitive than WT to binding sequence variations. In contrast, when comparing E80A HD with WT HD, many data points fell below the identity line and the relative binding energies regressed towards 0 on the E80A axis (Figures 1I and 1L). This suggests that E80A HD bound with lower specificity than WT HD and thus was more tolerant to base variations in the HD DNA motif. Different from R90W and E80A, K88N mutation dramatically altered CRX HD DNA-binding specificity (r: 0.160) (Figures 1J and 1M). The K88N preferred binding sequence (TAAT/ATT/A) is referred to as N88 HD motif hereafter. K88N HD also had the largest degree of discrimination from its preferred to the weakest binding motif, suggesting that it is most sensitive to variants in the HD DNA motif. As a control, we tested a second library with the TAANNN sites on the reverse strand and obtained similar results (Figures S2E-S2L). Together, these results indicate that while E80A mutation does not affect CRX HD DNA binding specificity, the K88N mutation dramatically alters the specificity in vitro.

E80A protein binds to WT sites while K88N occupies novel genomic regions with N88 HD motifs in vivo

Next, we asked if changes in DNA-binding specificity affected mutant CRX chromatin binding in developing photoreceptors. We first created two human mutation knock-in mouse models, CrxE80A and CrxK88N, each carrying a single base substitution at the endogenous Crx locus (Figures S3A and S3B, Methods). We confirmed that Crx mRNA was expressed at comparable levels in WT and mutant retinas (Figure S3C), and the full-length CRX proteins were readily detectable in the nuclear extracts from all samples (Figure S3D). We then obtained genome-wide binding profiles for each CRX variant by chromatin immunoprecipitation followed by sequencing (ChIP-seq) on mouse retinas at P14, a time when all retinal cell types are born, photoreceptor specification is completed in WT animals, and prior to any observed cell death in other CRX mutants previously characterized19,37. To focus on changes specific to each mutant CRX protein, only homozygous animals were used for ChIP-seq profiling.

Unsupervised clustering of all CRX binding sites revealed two major clusters (Figure 2A, Methods). Cluster 1 consisted of canonical WT CRX binding sites that are also occupied by CRX E80A protein (Figures 2A and 2B). Similar to WT CRX, CRX E80A binding in vivo was mostly enriched in intronic, followed by intergenic and TSS regions (Figure 3D). In contrast, CRX R90W, that also showed similar consensus preference to WT in vitro, failed to produce significant DNA binding in vivo (Figures 2A-2C, Methods). This suggests that the retinopathy phenotype of CrxR90W/W is likely due to loss of binding at canonical WT CRX binding sites. Intriguingly, while CRX K88N showed loss of binding at canonical CRX binding sites, it gained a small set of binding sites (Cluster 2, Figures 2A-2C). De novo motif searching with DREME38 under CRX peaks in each genotype revealed enrichment of monomeric HD motifs (Figure 2E) consistent with those found in Spec-seq (Figures 1G and 1K-1M), highlighted by a change in enriched HD motif from WT CRX HD type to N88 HD type in the CrxK88N/N retinas. Consistency with in vitro binding models suggests that in vivo changes in CRX chromatin binding were at least in part driven by the intrinsic changes in HD-DNA binding specificity by each individual mutation.

CRX E80A binds to WT sites while CRX K88N occupies novel genomic regions enriched for N88 HD motif in vivo.

(A) Enrichment heat map depicting CRX ChIP-seq normalized reads centered at all possible CRX peaks ± 2kb, sorted by binding intensity in WT samples. Clusters were defined by hierarchical clustering of CRX binding intensity matrix from all genotypes (STAR Methods). (B-C) Genome browser representations of ChIP-seq normalized reads for different CRX species in P14 WT and mutant mouse retinas at Rho and Atf2. (D) Enrichment heatmap showing fraction of CRX ChIP-seq peaks fall in different genomic environments. (E) Logo representations of de novo found short HD motifs under CRX ChIP-seq peaks in WT and mutant mouse retinas with DREME E-value on the righ

CRX-dependent activated genes affected in opposite directions in developing CrxE80A and CrxK88N mutant retinas

(A) Heat map showing sample-wise Pearson correlations of the expression of all CRX-dependent activated genes between P10 WT and HD mutant mouse retinas in this study (rows) with post-natal WT retinas from age P3 to P21 (columns, data from GSE87064). (B) Heat map showing the expression changes of DEGs in CRX-dependent activated gene set in HD mutant mouse retinas at P10. (C-D) Heat maps showing expression changes of selected photoreceptor genes from Group 1 and Group 2. Color scale identical to (B).

E80A and K88N mutations affected the expression of CRX-dependent activated genes in opposite directions in a critical time window of photoreceptor differentiation

To understand how different CRX mutations affected gene expression at canonical and de novo binding sites and how these changes impair photoreceptor differentiation, we turned to bulk RNA-seq analysis from the developing retinas at P10. At P10, photoreceptors have started to differentiate, and the expression of many photoreceptor genes undergo exponential increase39,40. To focus on the most relevant expression changes, we first defined a set of genes that most likely depend on CRX activity nearby for expression (Figure S4A, Methods). Briefly, we associated each CRX ChIP-seq peak to the nearest gene, filtered only genes with nearby CRX ChIP-seq peaks, and further narrowed the list of genes to those significantly down-regulated in the loss of function mutant CrxR90W/W. Gene ontology (GO) analysis confirmed that this putative CRX-dependent gene set is associated with biological processes related to photoreceptor development and functions (Figure S4B). This set of putative CRX-dependent genes also showed developmental dependent gain in expression, consistent with CRX’s primary function as a transcriptional activator (Figure S4C). As a control, CRX-independent genes were constitutively expressed and largely involved in general cellular processes (Figures S4D-S4F). Therefore, the CRX-dependent gene set comprises genes important for photoreceptor differentiation and functional maturation and are dependent on CRX for activation. We denote these genes as “CRX-dependent activated genes” (CRX-DAGs).

Next, we sought to understand how each mutation affected photoreceptor differentiation. One way of measuring the progression of photoreceptor differentiation is to determine the similarity in CRX-DAG expression in a given sample with that of known developmental ages in WT control animals. We thus performed sample-wise correlation of CRX-DAG expression obtained in our P10 samples with a previously published RNA-seq dataset of normal mouse retinal development (Figure 3A)39. As expected, our P10 WT sample showed strong correlations with all developmental ages in the published WT control dataset. A stronger correlation with early ages (P3, P7, P10) and a weaker correlation with later ages (P14, P21) is also an indication of ongoing photoreceptor differentiation at P10. Unlike the WT sample, CrxE80A/+ and CrxE80A/Asamples both showed a stronger correlation with later developmental ages (P14, P21) but a weaker correlation with earlier postnatal ages (P3, P7). Since the CRX-DAGs are normally developmentally upregulated, this shift in correlation towards later developmental ages suggested that these genes were prematurely upregulated in the P10 CrxE80A mutant retinas. In contrast, CrxK88N/+ and CrxK88N/N samples both showed a weaker correlation with all developmental ages when compared with WT samples in our dataset. This suggests that early photoreceptor differentiation was compromised in both CrxK88N mutants, consistent with their association with early-onset LCA23. Importantly, CrxR90W/W, also associated with LCA-like phenotype19,22, displayed strong correlation with earlier ages (P3, P7) similar to WT, but only showed moderate correlation with later ages (P14, P21). This suggests loss of CRX function at canonical binding sites does not affect the initiation of photoreceptor differentiation, but WT CRX activity at these sites is required to sustain differentiation. Since CRX-DAG expression was more severely affected in CrxK88N mutants than in CrxR90W/W, the photoreceptor differentiation deficits seen in the CrxK88N mutants cannot be explained solely by the loss of regulatory activity at canonical CRX binding sites. Overall, our sample-wise correlation analysis with normal retinal development dataset suggests that E80A and K88N mutations affected the expression CRX-dependent activated genes in opposite directions, implicating novel and distinct pathogenic mechanisms from the loss-of-function R90W mutation.

CrxE80A retinas show up-regulation of rod genes but down-regulation of cone genes, underlying CoRD-like phenotype

Upon closer examination, we noted that not all CRX-DAGs were up-regulated in CrxE80A mutants (Figure 3B). Hierarchical clustering of all CRX-DAGs using expression changes revealed two major groups (Methods). In aggregate, when compared to WT, Group 1 genes were up-regulated in CrxE80A mutants while Group 2 genes were down-regulated. We noted genes indicative of the two photoreceptor subtypes, rods and cones, could partially define the two groups (Figures 3C and 3D). For example, Esrrb41 and Nrl42 in Group 1 are important regulators of rod differentiation. Other genes in Group 1 are components of the phototransduction cascade in rods, including Rcvrn43, Rho44, Gnat145,46, Pde6g47, Abca448,49, Gnb150, Rdh1251, Cngb152, Rp153,54. Mis-regulations of many of these genes have been associated with diseases that affect rod development, function, and long-term survival. The increased activation of these genes likely underlies the stronger correlation with later developmental ages in CrxE80A+/and CrxE80A/A retinas (Figure 3A). In contrast, Group 2 genes, many down-regulated in CrxE80A mutants, were implicated in cone development and functions. For example, Gnat255,56, Pde6c57, and Pde6h58,59 all act in the cone phototransduction cascade. Mis-regulation of these genes has also been implicated in different retinal dystrophies that primarily affect cone photoreceptors. Comparison of ChIP-seq signal revealed that peaks associated with Group 2 genes showed lower occupancy compared to Group 1 genes in the CrxE80A/A retinas (Mann-Whitney U test p-value: 9.51e-07) while no difference was observed in WT retinas (Mann-Whitney U test p-value: 0.541), suggesting loss of CRX activity likely underlies the down-regulation of Group 2 genes in the CrxE80A mutants (Figure S5A). Collectively, the selective down-regulation of cone genes in Group 2 may explain the CoRD-like phenotype in adult CrxE80Amutant mice described later.

Additionally, we noticed that a subset of genes not affected in CrxR90W/W(CRX-independent genes) were also down-regulated in CrxE80A mutants (Figures S5B-S5D). Among these genes were transcription regulators important for early photoreceptor development, such as Ascl160, Rax61, Sall362, and Pias363. The down-regulation of these factors coincided with the up-regulation of mature rod genes in P10 CrxE80A retinas, suggesting that the E80A mutation might hamper the proper timing of photoreceptor differentiation.

CrxK88N retinas display greater reduction of rod and cone genes than the loss-of-function mutants

CrxK88N retinas had the most severe gene expression changes among all mutants with down-regulation of both Group 1 (rod) and Group 2 (cone) genes (Figures 3B-3D). The heterozygous CrxK88N/+ retina displayed a similar degree of expression reduction as homozygous CrxR90W/W, consistent with its association with dLCA23. Given the normal phenotype of heterozygous loss-of-function mutants - Crx+/-and CrxR90W/+19, these results suggest that mutant CRX K88N not only failed to activate WT target genes, but also functionally antagonized WT CRX regulatory activity in differentiating photoreceptors. This antagonism might be associated with ectopic CRX K88N activity when bound to regulatory regions with N88 HD DNA motifs (Figures 2A and 2E). In the absence of WT CRX, CrxK88N/N retina displayed a more severe expression reduction of CRX-DAG than CrxR90W/W, raising the possibility that CRX K88N also antagonized the activity of other transcriptional regulators important for photoreceptor differentiation. Supporting this possibility, a set of CRX-independent genes were also mis-regulated in both heterozygous and homozygous CrxK88N mutants. We noted that a number of these down-regulated genes are also involved in photoreceptor functional development (Figures S5B, S5E, and S5F). Overall, CRX K88N is associated with greater gene expression changes than other CRX HD mutants, which is likely attributed to ectopic regulatory activity.

Both E80A and K88N mutants show compromised rod/cone terminal differentiation in young adults

Since Crx+/- and CrxR90W/+ mutant mouse models showed a late-time recovery in photoreceptor gene expression and function20, we sought to determine the degree of photoreceptor differentiation in CrxE80A and CrxK88N mutants at P21. At this age, the normal retina has largely completed terminal differentiation with photoreceptor gene expression reaching a plateau64. However, when P21 CRX HD mutant retinas were examined for the expression of genes under the GO term detection of light stimulus (GO:0009583), which comprises genes in both rod and cone phototransduction cascades, many genes failed to reach WT levels, despite variable degrees of impact across different HD mutants (Figure 4A). CrxE80A mutants, in contrast to the increased rod gene expression at P10, displayed a deficiency in both cone and rod phototransduction genes at P21 (Figure 4B, Supplementary Table 6). This suggests that mutant CRX E80A transcriptional activity fails to sustain photoreceptor terminal differentiation and ultimately results in non-functional and severely affected photoreceptors. In comparison, CrxK88N mutants showed severely reduced expression of rod/cone phototransduction genes at both P10 and P21 (Figures 3B-3D and 4B).

Photoreceptor genes important for phototransduction are down-regulated in all HD mutants

(A) Box plot showing that genes in the detection of light stimulus GO term were down-regulated and affected to various degrees in different adult (P21) HD mutant mouse retinas. (B) Heat map showing that expression of both cone and rod phototransduction genes were down-regulated in adult (P21) HD mutant mouse retinas. Annotation of rod and cone enrichment of each gene is in Supplementary Table S6. See Supplementary Figure S6 for the developmental expression dynamics of these genes.

To assess retinal morphology and photoreceptor subtype-specific gene expression at the cellular level, we performed immunohistochemistry analysis on P21 retinal sections. In WT animals, a hallmark of photoreceptor maturation is the outgrowth of photoreceptor outer segments (OS) filled with proteins necessary for the phototransduction. We thus performed hematoxylin and eosin (H&E) staining on P21 sagittal retinal sections to visualize changes in retinal layer organization, focusing on photoreceptor layers – outer nuclear layer (ONL) and OS. Compared to the well-organized ONL in WT retinas (Figure 5A), all mutants showed variable degrees of ONL disorganization, forming waves, whorls, and rosettes (Figures 5B-5E). The ONL disorganization was more severe in homozygotes than in heterozygotes for both mutations and CrxK88Nmutants were more severely affected than CrxE80A mutants. Photoreceptor OS layer was formed in the CrxE80A/+ retinas, but absent in CrxE80A/A, CrxK88N/+, and CrxK88N/N mutant retinas. Inner retinal layers, including inner plexiform layer (IPL) and ganglion cell layer (GCL) were not as severely affected as the outer retinal layers, supporting a model that the mutant morphological abnormalities largely originated from the diseased photoreceptors. These morphological abnormalities were distinct from the degenerative phenotypes of other Crx mutant models reported previously19,65,66.

Only CrxE80A/+ retinas maintain photoreceptor OS and residual rod ERG response

(A-E) Hematoxylin-eosin (H&E) staining of P21 retina sections show that photoreceptor OS layer is absent in all mutant retinas except CrxE80A/+. OS: outer segment; ONL: outer nuclear layer; INL: inner nuclear layer; IPL: inner plexiform layer; GCL: ganglion cell layer. Scale bar, 100µm. (F-J) Rhodopsin (RHO, red) immunostaining is present in CrxE80A/+, CrxE80A/A, and CrxK88N/+ retinas and absent in CrxK88N/N retina. Cone arrestin (mCAR, green) immunostaining is absent in all mutant retinas. Nuclei were visualized by DAPI staining (Blue). Scale bar, 100µm. (K-M) The electroretinogram responses (ERG) recorded from 1-month mice. Error bars represent the standard error of the mean (SEM, n ≥ 4). p-value: Two-way ANOVA and Tukey’s multiple comparisons. ****: p ≤ 0.0001. ns: >0.05.

Next, we sought to determine the expression of the rod-specific visual pigment rhodopsin (RHO) and cone arrestin (mCAR) in the P21 mouse retinas. In WT retinas, RHO is trafficked to the rod OS while mCAR is present in the cone OS and IS (inner segment), cell body, and synaptic terminals (Figure 5F). Unlike WT retina, all mutants lacked mCAR immunoreactivity (Figures 5G-5J), consistent with the loss of cone gene expression shown by RNAseq (Figure 4B). In CrxE80A/+ retinas, RHO staining was localized to the OS layer; in CrxE80A/A and CrxK88N/+retinas, positive RHO staining was observed within the whorls and rosettes; in CrxK88N/N mutant retinas, RHO staining was completely absent. Importantly, we did not observe mis-localized RHO staining in the inner retinal layers (INL) suggesting that the developmental programs of other retinal cell types were not directly affected by E80A or K88N mutation. Overall, abnormalities in the cone/rod gene expression matched the corresponding human disease diagnosis21,23, and the phenotypic severity correlated with the degree of mis-regulation of CRX target genes in the corresponding RNAseq dataset. Thus, these results support a model that CRX HD mutation-mediated mis-regulation of gene expression disrupts photoreceptor terminal differentiation and leads to defects in retinal layer organization and OS formation.

E80A and K88N mouse models show visual function deficits that recapitulate human diseases

To understand the consequences of disrupted photoreceptor differentiation on visual function, we measured electroretinogram (ERG) responses to light stimuli for WT and mutant mice at one month of age (Figures 5K-M). In response to incremental changes of light intensities, WT animals showed corresponding amplitude increases in dark adapted A-waves (rod signals) and B-waves (rod-evoked bipolar cell signals), as well as in light-adapted B-waves (cone-evoked bipolar cell signals). The three severe mutants, CrxE80A/A, CrxK88N/+, and CrxK88N/N had no detectable dark-adapted or light-adapted ERG responses, suggesting that these mice have no rod or cone function and are blind at young ages. The null ERG phenotype of the CrxK88N animals is consistent with the clinical LCA phenotype in humans23. In contrast, CrxE80A/+ animals retained partial rod ERG responses as indicated by the reduced A-wave and B-wave amplitudes (Figures 5K and 5L). Yet, CrxE80A/+ animals had no detectable cone ERG responses, which is consistent with the CoRD clinical phenotype in humans21 (Figure 5M). Taken together, the visual function impairment in each CRX HD mutant model, coincided with the morphological and molecular changes, suggesting that CrxE80A and CrxK88Nmouse models recapitulate the corresponding human diseases.

CRX E80A has increased transactivation activity and leads to precocious differentiation in CrxE80A retinas

Lastly, we asked what might be the molecular mechanism that causes the mis-regulation of photoreceptor genes in the mutant retinas. Previous studies have established that reporter assays with the rhodopsin promoter in HEK293T cells can measure changes of CRX transactivation activity and inform the mechanisms by which photoreceptor genes are mis-regulated in CRX mutant retinas18. We thus tested the transactivation activity of the three CRX HD mutants on the pRho-Luc reporter in HEK293T cells (Figure 6A, Methods). Consistent with published studies, R90W mutant had significantly reduced activity compared to WT CRX18. K88N mutant showed a similarly reduced activity as R90W, consistent with K88N’s loss of binding at canonical CRX sites and failure to activate CRX-DAGs in vivo (Figures 2A, 2B, 3A-3C). In contrast, E80A mutant, which binds to canonical CRX sties, showed significantly increased transactivation activity on the Rho promoter. This hyperactivity of E80A protein at Rho promoter correlates with the upregulation of CRX-DAGs in the mutant retinas at P10 (Figures 3A-3C).

CRX E80A hyperactivity underlies precocious photoreceptor differentiation in CrxE80A retinas.

(A) Boxplot showing luciferase reporter activities of different CRX variants. P-values for one-way ANOVA with Turkey honestly significant difference (HSD) test are indicated. p-value: ****: ≤0.0001, ***: ≤0.001, ns: >0.05. (B-D) Rhodopsin (RHO, green) immunostaining is absent in P3 WT retina but detected in CrxE80A/+ and CrxE80A/A retinas. Nuclei are visualized by DAPI staining (Blue). Arrow indicates the sporadic RHO staining in CrxE80A/+ sample. ONBL: outer neuroblast layer; GCL: ganglion cell layer. Scale bar, 100µm.

A transition to the next developmental stage usually requires the expression of important developmental genes passing an abundance threshold. Based on the hyperactivity model, photoreceptor genes are activated stronger in CrxE80A retinas and thus could reach the abundance threshold earlier. To determine the consequences of E80A hyperactivity on photoreceptor differentiation timing, we compared RHO protein expression during early postnatal retinal development using retinal section immunostaining (Figures 6B-6D). In WT retinas, most rods were born by P3 but had not differentiated (Figure 6B). Previous studies showed that RHO proteins were detected by IHC starting around P7 in WT retinas64. In comparison, both CrxE80A/+ and CrxE80A/A retinas showed positive RHO staining at P3 (Figures 6C and 6D). RHO+ cells were largely seen in the outer portion of the ONBL layers in CrxE80A/+ retinas, and strikingly spread throughout the large presumptive ONL layers in CrxE80A/A retinas. The detection of RHO protein in P3 CrxE80A mutant retinas indicates that photoreceptor differentiation program was precociously activated. Taken together, our results support a model that E80A and K88N mutations each perturbs CRX regulatory activity in a unique way, causes photoreceptor differentiation defects, and ultimately leads to distinct dominant disease phenotype that recapitulates human diseases.

Discussion

Through molecular characterization of mutant proteins, transcriptome, and cellular profiling of developing mutant mouse retinas and ERG testing of adult retinas, we have identified two novel pathogenic mechanisms of CRX HD mutations, E80A and K88N, that are associated with dominant CoRD and dominant LCA in human21,22. Distinct from the previously characterized loss-of-function R90W mutation19,20, E80A and K88N mutations produce altered CRX proteins with gain of regulatory functions - CRX E80A is associated with increased transcriptional activity and CRX K88N has altered DNA binding specificity. Both CRX E80A and CRX K88N proteins impair photoreceptor gene expression, development and produce structural and functional deficits in knock-in mouse models, recapitulating human diseases65 (Figure 7). Thus, both target specificity and regulatory activity precision at the canonical CRX targets are essential for proper photoreceptor development and functional maturation.

Missense mutations in CRX HD affect photoreceptor gene expression and leads to distinct retinal disease phenotypes through gain- and loss-of-function mechanisms.

Although associated with distinct disease phenotypes, the E80A, K88N, and R90W mutations are located very close to each other in the CRX HD recognition helix. Extensive biochemical and structural studies on the HD-DNA complexes afford important insights into how these mutations could affect CRX HD-DNA interactions differently. Most distinctively, CRX K88 residue, at HD50 position, is the major contributor to HD-DNA binding specificity, with lysine making favorable interactions with both guanines in the CRX consensus TAATCC binding site67,68. It is thus expected that K88N mutation drastically changes CRX DNA binding preference at the 3’ end of the HD motif, reminiscent of previous findings on novel HD DNA-binding specificity using bacterial one-hybrid (B1H) systems31,69. Supporting the importance of K88 residue mediated CRX target specificity in regulating photoreceptor development, CrxK88Nretinas show more severe perturbations in photoreceptor gene expression and development than in loss-of-function mutant CrxR90W/W. CrxK88N/+and CrxK88N/N mice display the most severe photoreceptor morphological deficits observed in any Crx mouse models and show absence of visual functions in young adults. Thus, CRX target specificity is critical for photoreceptor development fidelity.

In developing WT mouse retinas, the HD motif preferred by HD TFs with a glutamine (Q) at HD50 position encodes quantitatively different activity than the CRX consensus suggesting functional difference between HD binding site variants61,70. It is likely that the severe CrxK88N phenotypes are attributed to both diminished activity at canonical CRX motifs and ectopic binding and transcriptional activity at N88 HD motifs. Since a functional copy of WT CRX is retained in CrxK88N/+ retinas, the lack of WT activity alone cannot explain the severe developmental deficits. Alternatively, these results suggest involvement of additional regulatory mechanisms: CRX K88N activity at N88 HD motifs might (1) ectopically activate genes whose expression prevents the progression of development or inactivate genes required for development; (2) interfere with other HD TFs that also recognize N88 HD motifs; (3) lead to epigenetic alterations that antagonize normal CRX functions. Many other HD containing TFs are expressed in developing mouse retina, including OTX2, RAX, VSX2, PAX6, SIX3/6, and LHX family6. Different from CRX, these HD TFs are essential for gene regulation in retinal progenitor cells and/or in other retinal cell lineages. Alteration of CRX DNA binding specificity could mis-regulate genes originally targeted by these HD TFs and lead to severe perturbations in the retinal gene regulatory networks (GRNs). To date, most studies have focused on CRX activity at cis-regulatory sequences enriched for the WT CRX consensus motifs71,72. Systematic comparison of regulatory activity at N88 HD motifs and WT consensus in the context of photoreceptor development in both WT and mutant retinas would be needed to substantiate the impact of mutant CRX K88N activity at different HD motifs. These experiments will also help clarify the pathogenic mechanisms in the CrxK88Nmodels and extend our knowledge of CRX HD mediated regulatory grammar during photoreceptor development.

Different from CRX K88, the E80 and R90 residues, although the most common residues at HD42 and HD52 positions respectively, do not contact DNA directly and thus lacked in-depth investigations in prior studies. CRX R90 residue has been suggested to confer additional stability for the homeodomain fold besides the core residues and make contacts with the DNA backbone through bases in the TAAT core motif67,68. Substitution of the basic R90 residue with a bulky, neutral tryptophan (W) potentially reduces overall CRX HD stability which in term reduces CRX HD DNA-binding affinity without affecting its binding preference. The potential reduction in CRX HD-DNA complex stability is in line with our observation that R90W mutation abolishes CRX binding across the genome resulting in global loss of CRX target gene activation. It also explains the association of CRX R90W mutation with recessive loss-of-function LCA phenotypes in human and mouse22,65.

Structural studies suggest that CRX E80 residue plays a role in stabilizing the HD-DNA binding complex through intramolecular interactions with other HD residues29,68, yet functional validations await further experiments. E80A mutation, replacing glutamic acid (E), which is acidic and polar, with alanine (A), which is neutral and non-polar, could render HD-DNA interactions more promiscuous as reflected in overall reduced magnitude of CRX E80A HD specificity (Figure 1I). Regulatory sequences of many photoreceptor genes contain both consensus and non-consensus CRX motifs14,15,18,35. It is likely that the more promiscuous CRX E80A-DNA interaction increases the likelihood of non-consensus CRX HD motifs being bound and activated resulting in overall increased transcriptional output (hyperactivity) as seen both in luciferase assays and in developing CrxE80A mouse retinas. The promiscuous TF-DNA-binding associated hyperactivity phenomenon has also been observed in a dominant disease mouse model harboring a missense mutation in the zinc finger TF Krüppel-like factor-1 (KLF1)73. Yet, in adult CrxE80A retinas, photoreceptor terminal differentiation is impaired, resulting in disrupted retinal morphology and defective visual functions. Photoreceptor differentiation is programmed via sequential and concerted gene expression programs within a defined time window74,75. One explanation for adult CrxE80A phenotypes is that CRX E80A hyperactivity precociously activates later stage genes in the absence of proper nuclear context and/or subcellular structures, which in turn negatively impacts early events in photoreceptor differentiation. These observations underscore the importance of precisely tuned CRX-mediated transcriptional activity during photoreceptor development.

Although a global increase in expression is expected in CrxE80A retinas based on the hyperactivity model, a subset of CRX dependent activated genes implicated in cone photoreceptor development and functions is down-regulated in both differentiating and mature mutant retinas. While cones undergo terminal differentiation to develop into cone subtypes – M- or S-cones – in a similar postnatal window as rods, they were born prenatally in mice within an earlier time window than rods. At an early postnatal age, cells expressing RXRγ, a ligand-dependent nuclear hormone receptor normally expressed in developing cones60,76–78, were observed in CrxE80A retinas, suggesting cone photoreceptors were born in these mutant retinas (Figure S7A). One model for lack of cone markers in adult CrxE80A retinas is that CRX E80A improperly activates later stage cone genes at a much earlier time window, disrupting cone terminal differentiation. Supporting this model, CRX E80A also hyperactivates the S-cone opsin promoter (Figure S7B). An alternative model is that cones might be more sensitive to perturbations in CRX activity. It is known that cones depend on a different repertoire of TFs than rods for subtype terminal differentiation60. It is possible that cone TFs respond differently to mutant CRX E80A hyperactivity, leading to the distinct expression changes in CrxE80A mutant retains. It is important to note that different point mutations at CRX E80 residue have been reported in dominant CoRD cases (ClinVar VCV000865803.1, VCV000007416.7, VCV000099599.6), emphasizing the importance of residue CRX E80 in regulating cone photoreceptor development. Since cones only make up a very small portion (3%) of photoreceptors in mouse retinas79, quantitative characterization of CRX E80A molecular functions in a cone dominant retina warrants further study to understand its selective effect on the cone differentiation program and help elucidate WT CRX regulatory principles in early photoreceptor development.

Given that the spatial structures and HD-DNA contact models of HD proteins are evolutionarily conserved, our study of CRX provides valuable molecular insights for HD mutations implicated in other diseases. For example, p.E79K substitution (corresponds to CRX E80) in OTX2 HD is associated with dominant early-onset retinal dystrophy80, heterozygous p.R89G (corresponds to CRX R90) mutation in OTX2 HD causes severe ocular malformations81, and missense mutations of the CRX K88 and R90 homologous residues in PITX2 HD are associated with dominant Rieger syndrome82. It is likely that these mutations affect HD activity in similar ways as observed in CRX, and the exact disease manifestation is determined by cell-type or tissue specific mechanisms. The retina is readily accessible, and a broad range of molecular tools are available for ex vivo and in vivo manipulations. We believe that CRX is an ideal model to study the pathogenic mechanisms of HD mutations and to test therapeutic regimens, which would ultimately benefit the study of HD TFs and their associated diseases in other tissues and organs.

One limitation of this work is that effects of E80A and K88N mutations on CRX HD-DNA interactions have been evaluated at monomeric HD motifs and with homogenous protein species both in vitro and in vivo. Further evaluation of WT and mutant CRX binding at dimeric motifs will be desirable, since selected dimeric HD motifs are known to mediate HD TF interactions to ensure gene expression fidelity during development83,84. Relatedly, we also need to address how CRX WT and mutant E80A or K88N proteins interact at HD binding motifs – whether they cooperate or compete with each other, whether these interactions are HD motif sequence-dependent, and how gene expression is impacted by CRX cooperativity or competition. While CRX HD mediates both TF-DNA interactions and protein-protein interactions, evaluation of how E80A and K88N mutations impact CRX interaction with other important photoreceptor TFs and how perturbations in these interactions lead to disease phenotypes warrant further study.

Collectively, our findings support a unifying model in which precise CRX interaction with cis-regulatory sequences is essential for gene expression and functional maturation during photoreceptor development. Disease-associated mutations in CRX have been classified into two main groups – insertion/deletion-derived frameshift mutations in the activation domain (AD) and missense mutations in the homeodomain65. Prior biochemical and mouse model studies of the first group have established that AD-truncated mutant proteins abolish CRX transcriptional activity and functionally interfere with the WT allele. As a result, the mutant retinas fail to activate or maintain robust cone/rod gene expression, resulting in incomplete photoreceptor differentiation and ultimately rapid degeneration of immature photoreceptors16,19. In this study, we demonstrate that missense mutations in the CRX HD, by either a loss- or gain-of-function mechanism, alter CRX target specificity and/or CRX transactivation activity. These biochemical property changes impair CRX-mediated transcriptional regulation in vivo and lead to distinct morphological and functional deficits (Figures 7A-7D). Despite the difference in molecular mechanisms, both CrxE80A and CrxK88N mouse models develop whorls and rosettes in the ONL by P21, which are not observed in degenerative CRX mouse models19, suggesting distinct pathogenic mechanisms. Future cellular biology studies are needed to understand the formation mechanisms of these unique cellular phenotypes (ONL disorganization) and their impacts on the function and survival of photoreceptors and other retinal cell types over development.

Our study here also emphasizes the importance of tailoring gene therapy regimens to tackle individual pathogenic mechanisms. For instance, while supplementing WT CRX might be sufficient to rescue a hypomorphic/loss-of-function mutant, simultaneous elimination of a gain-of-function CRX product would be necessary to rescue dominant mutants, as exemplified in a recent report of allele-specific gene editing to rescue dominant CRX-associated LCA7 phenotypes in a retinal organoid model85. We believe that this principle also applies to other dominant neurological diseases. Additionally, with the refinement of the CRX mechanistic model, when new disease mutations are identified, genetic counsellors can now provide more informed predictions of disease progression and future visual deficits. This information is important for individuals to be psychologically prepared and seek necessary assistance to improve their quality of life.

Supporting information